### 2022-06-24 Readme.Rmd
first commit

---
title: Identifying covariate traits of trance and possession phenomena in the Ethnographic
  Atlas using sparse regression
author: "Péter Rácz"
date: "`r format(Sys.Date(),'%e %B, %Y')`"
output:
  md_document:
    variant: markdown_github
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, fig.path = 'figures/', fig.width = 8, fig.height = 8)
```

```{r header}
setwd('~/Github/Racz2023trance/')

library(tidyverse)
library(glue)
library(knitr)
library(patchwork)

d = read_tsv('data/dat_long.tsv')
sccs = filter(d, in_sccs)
```

We want to look at two questions. What sort of social factors have been proposed to cooccur with trance and possession phenomena? Using the Ethnographic Atlas and the Standard Cross-Cultural Sample, can we get more confident that these factors do cooccur with such phenomena across cultures?

## 1. Trance and possession phenomena have societal covariates in the literature

Many people have argued that trance induced by possession is not a random, idiosynchratic pathology but a recurrent psychosocial trait that is at least partly motivated by other social factors.

High overall social complexity and social rigidity -- the resulting power dynamics -- might translate to possession trance practices. Such practices might serve to express or relay perceived social inflexibility and inequality. This is echoed in the medical anthropology literature (@during2011critical,@dein2012religion) and ties in with broader questions of inequality and mental health (@nguyen2003anthropology). Various causative models have been proposed to account for the emergence of possession trance and, specifically, shamanism in a social setting (@winkelman2015shamanism, @singh2018cultural, @wood2018trance).

Some or most groups can feel stuck in a society. If this makes trance and possession more likely, we should see correlations between measures of social complexity, organisation, and inequality and possession and trance in cross-cultural data. Many such correlations have been proposed by previous work. Here is a list, cross-referenced with @wood2018complexity:

| Relationship                                                                             | Reference                |
|:-----------------------------------|:-----------------------------------|
| less complexity $\rightarrow{}$ trance w/o possession                                    | @bourguignon1968         |
| women's social participation                                                             | @bourguignon1973religion |
| increased role differentiation and slavery parctices $\rightarrow$ trance and possession | @greenbaum1973           |
| main mode of subsistence                                                                 | @bourgouignon1977        |
| increased social participation $\rightarrow{}$ possession                                | @swanson1978             |
| more complexity $\rightarrow{}$ trance w/ possession                                     | @winkelman1986           |
| cognatic kinship $\rightarrow{}$ more possession and trance                              | @shaara1992              |
| decreased social participation $\rightarrow{}$ possession                                | @douglas2004natural      |
| monotheism                                                                               | @wood2018complexity      |

Several works made the same point or emphasised various aspects of the same relationship. The above table is a bit reductive.

It's an entirely different question whether we can build more ambitious explanatory models that link possession trance and similar practices to e.g. social inequality. This is not a question we really want to deal with here. What we want to know is whether the above descriptions are accurate at all, when given closer scrutiny. We can answer this question by looking at a cross-cultural data set, the Ethnographic Atlas, to see if we can attest in it any of the correlations proposed in the literature.

## 2. This can be mapped to the Ethnographic Atlas

We find data on trance and possession phenomena and their various proposed covariates in the Ethnographic Atlas [@murdock1967ethnographic] (we use the version made available in the D-Place database, @kirby2016d).

The Atlas records the presence of trance and possession phenomena in a given society, using the following variable (EA112):

```{r }
d %>% 
  filter(var_title == 'Trance states', !is.na(code)) %>% 
  distinct(code, var_description) %>% 
  arrange(code) %>% 
  rename(coding = code, Description = var_description) %>% 
  kable('simple')
```

This is basically a truth table of whether possession beliefs or trance states are present in a given society, with some causal links between the two (if possession happens it leads to trance, if trance happens it's because of possession, etc). We can think of different groupings (trance phenomena encompass levels 1,3,4,5,6,7, possession phenomena encompass 2,3,4,5,6,7, specifically possession trance covers 3,4,6,7 but not 1,2,5,8).

The eight levels here likely result from a trade-off between description accuracy and generalisability that is present throughout the Atlas. It is unlikely that eight descriptions suffice to capture the plethora of phenomena related to trance and possession. But largely the same is true for all variables in the Atlas. This hasn't stopped people from doing large-scale comparative studies before!

The Atlas lists a set of variables that tie in with the observed covariates in the literature. Here are some relevant-looking ones:

```{r }
d %>% 
  filter(var_title != 'Trance states') %>% 
  distinct(var_id,var_title,var_definition) %>%
  arrange(var_id) %>% 
  rename(ID = var_id, Name = var_title, Definition = var_definition) %>% 
  kable('simple')
```

The covariates in the literature do not directly map on the variables in the Atlas. The variables are themselves correlated, either because they measure different facets of the same thing (like how complex a society is) or because they were adapted together by socially or geographically close societies (Galton's problem).

## 3. How do we look for correlations between trance and possession and their proposed societal covariates?

We could table how often trance and possession phenomena and their posited covariates pattern together. This is generally a bad idea.

There are eight types of trance and possession phenomena recorded in the Atlas (and we can make up new codings!) and a wide set of possible covariates, all with a number of levels:

```{r }
vars = d %>% 
  filter(var_title != 'Trance states') %>% 
  distinct(var_id,var_title,var_name) %>%
  arrange(var_id) %>% 
  group_by(var_id, var_title) %>% 
  nest() %>% 
  mutate(levels = map(data, ~ paste(.$var_name, collapse = ', '))) %>% 
  select(var_id, var_title, levels) %>% 
  unnest(cols = c(levels)) %>% 
  rename(ID = var_id, Name = var_title)

kable(vars, 'simple')
```

Even if we initially restrict ourselves by a hypothesis, if we keep trying all the possible combinations, we are bound to find something. In addition, data are missing from the Atlas in a non-random manner (e.g. if a society is hard to access, it is harder to provide a population size estimate for it). This makes any catch even more spurious.

We could assess the probability of all eight descriptions in the Atlas together by using a multinomial model. But because data are missing for the covariates and because the covariates are themselves correlated, a single model will be wasteful (losing all the observations for which data are missing) and badly fit (with a lot of colinear predictors in it). We're probably better off building a lot of models and then aggregating over them to find the most important covariates of trance and possession phenomena.

We can do this by using an ensemble learning method. This means that we start with societies with information on trance and possession phenomena. We build a lot of tiny models, using random samples of the data and random subsets of our covariates. Then, we compare these to assess the overall importance of our covariates. Building little models in parallel is called bagging. This is how a random forest works, growing many classification or regression trees and aggregating over the results. One step further is to train little models on the mistakes of the previous little models. This is called [boosting](https://www.ibm.com/cloud/learn/boosting) and usually results in higher training and prediction accuracy.

## 4. Building a boosting model

In this analysis, we will use a gradient boosting model on the data from the Ethnographic Atlas, subsetting it to only include societies in the Standard Cross-Cultural Sample.

Gradient boosting is an ensemble learning method which uses decision trees. A decision tree is a directed tree graph in which each node represents a decision or split based on a predictor variable and each leaf is an individual data point. It can be used to classify leaves by grouping them under different node structures. A gradient boosting algorithm fits such trees in a sequence, training each on the errors of the previous one, minimising overall error. It stops when further iterations do not result in meaningful improvement on model accuracy (according to pre-set criteria) [@cook2016practical,@rhys2020machine,@smith2017decision].

We only fit the model on societies which (a) have information on whether trance and possession phenomena are present (we can't do much with those where we don't know this) and (b) come from the standard cross-cultural sample, which minimises the phylogenetic signal so we don't have to worry about it.

We use hyperparameter tuning to find the best possible model. The outcome of the model is one of the eight descriptions of trance and possession phenomena in the Atlas. The predictors are the possible covariates identified above. The model is fit on a training set of `length(unique(sccs$soc_id))` societies with information on trance and possession in the standard cross-cultural sample.

I'm scarce on the technical details here on purpose, but the analysis can be replicated with the code in this repository.

## 5. Results on the training sample

Here is the distribution of the various descriptions for trance and possession in our training sample. I gave the descriptions shorter names as well, to make things easier later on:

```{r }
sccs %>% 
  filter(var_id == 'EA112') %>% 
  count(var_name,var_description) %>% 
  rename(name = var_name, description = var_description, count = n) %>% 
  add_row(name = 'Total', description = '', count = 151) %>% 
  kable('simple')
```

We see that only about 10% of societies have no trance or possession phenomena. From the rest, a majority have possession trance in some form.

We compare this with how our model categorises each society. This is our model's confusion matrix. The rows are the actual descriptions, the columns are the predicted descriptions. So a good model would have large counts on the diagonal (where it got the label right) and low values elsewhere (where it didn't). The last two columns give the model's error as a probability (0-1) and as a rate (as n/m).

```{r }
cm1a = read_tsv('models/conf_matrix11model.tsv')
cm1a %>% 
  mutate(original = c(names(cm1a)[!names(cm1a) %in% c('Error','Rate')],'Total')) %>% 
  relocate(original) %>% 
  kable('simple', digits = 2)
```

These results are not very impressive. The model gets some of the larger categories less wrong but it is completely off the mark on the smaller ones. (It gets only one society right in the "no trance or possession" category and all of them wrong in the "possession-only" category!)

Focussing on the relationship between category sizes and model accuracy, we can tell two stories here. The first one is that the model is terrible and it is only likely to do better with the larger categories because of their size (so that the baseline probability of category membership is higher). The second story is largely the same but with some hopeful notes.

We can say that the way trance and possession phenomena are defined in the Atlas is very fragmentary. It involves at least two external sources of noise. First, for a category like "trance only, no possession", maybe the curators of the Atlas put very disparate psychosocial phenomena into this category, which haven't a lot to do with each other at all. In this case, it is no small wonder that we can't capture this category algorithmically. Maybe it's all arbitrary!

Second, maybe, even for the larger categories, the boundaries are porous. Maybe it's not entirely clear whether, in a given society, only possession explains trance phenomena. Maybe the fieldworkers missed something! Maybe they disagree on their definitions of things (e.g. some say that a trance phenomenon in a given society is not caused by possession ergo it is not trance at all but something else).

We can work around some of these issues by re-defining our outcome variable. We split our descriptions into two categories, based on whether they capture possession trance or not. This would look like this:

```{r }
d %>% 
  filter(var_title == 'Trance states') %>% 
  distinct(var_description) %>% 
  rename(description = var_description) %>% 
  add_column(`this is...` = c('something else','something else','possession trance','possession trance','possession trance','something else','something else','possession trance')) %>% 
  kable('simple')
```

Now we have two categories that are roughly the same size and are perhaps less conceptually ambiguous. We refit the model, but this time it only tries to guess whether a society has possession trance or something else (including no trance or possession at all).

## 6. Results on the training sample, part II

We look at the model's confusion matrix:

```{r }
cm4a = read_tsv('models/conf_matrix41model.tsv')
cm4a %>% 
  mutate(original = c(names(cm4a)[!names(cm4a) %in% c('Error','Rate')],'Total')) %>% 
  relocate(original) %>% 
  kable('simple', digits = 2)
```

This is a lot better. Granted, there are only two categories, but the error rate is much lower. This means our model can more or less use the societal covariants to guess whether a society, by the Atlas' own terms, will have possession trance phenomena (1) or not (0).

Having established this, we can ask the model to make predictions for the entire Ethnographic Atlas (where trance and possession data are available, of course) -- including the SCCS subset. Generally, you want to separate the training and test sets in Machine Learning and not do this, but the 'everything except the SCCS' subset of the Atlas would be an odd test set.

Here is the confusion matrix for all `r length(unique(d$soc_id))` societies in the Atlas where trance and possession information are available:

```{r }
cm4b = read_tsv('models/conf_matrix41test.tsv')

cm4b %>% 
  add_column(original = c(0,0,0,1,1,0,1,1)) %>% 
  group_by(original) %>% 
  summarise(`0` = sum(`0`), `1` = sum(`1`)) %>% 
  kable('simple')

# d %>% 
#   filter(var_id == 'EA112') %>% 
#   mutate(
#     possession_trance_present = case_when(
#       code %in% c(3,4,6,7) ~ T,
#       code %in% c(1,2,5,8) ~ F
#         )
#   ) %>% 
#   count(possession_trance_present) %>% 
#   kable('simple')
```

This is much worse than the training data only but that is at least partly because the model is ignorant of the phylogenetic signal which might result in patterns that are unexpected from a purely correlational point of view. And it is still not terrible! It gets 468/`r length(unique(d$soc_id))` societies right. There are 281 versus 377 societies in the data with and without possession trance. If it assumed that no societies had possession trance, it would get 377 data right, and 468 is considerably larger number than that. This means the model is doing something right.

Now that we established that the model does find some purchase, we take a look at the variables that were most important in predicting the presence of possession trance in a given society. I will only show societal covariants with a high enough importance:

```{r }
varimp4 = read_tsv('models/varimp41.tsv')
vars2 = varimp4 %>% 
  filter(relative_importance > 15) %>%
  mutate(ID = str_extract(variable, '^EA...')) %>% 
  left_join(vars, by = "ID") %>% 
  select(ID, Name, relative_importance)

vars2 %>% kable(digits = 2, 'simple')
```

What do we see? The five most important driving factors are the prevalence of slavery, family complexity, the complexity of class structure, premarital norms for women, and settlement patterns. Various aspects of social complexity and filial and consanguineal organisation also make the list. We can cross-reference this with our first table, with pointers in the literature. We add a column indicating whether the Atlas seems to support each specific claim. (This we do in a very haphazard manner!)

| Relationship|Reference|Our model|
|:-|:-|:-|
|less complexity$\rightarrow{}$trance w/o possession|@bourguignon1968|not directly, but more complexity means possession trance is more likely|
|women's social participation|@bourguignon1973religion|not directly, but premarital sexual norms can be a proxy|
|increased role differentiation and slavery practices$\rightarrow$trance and possession|@greenbaum1973|slavery, social and class complexity are relevant|
|main mode of subsistence|@bourgouignon1977|not really|
|increased social participation$\rightarrow{}$possession|@swanson1978|indirectly, through domestic and social complexity|
|more complexity$\rightarrow{}$trance w/ possession|@winkelman1986|yes, see above|
|cognatic kinship$\rightarrow{}$more possession and trance|@shaara1992|not directly, but family and kinship seems to do something|
|decreased social participation$\rightarrow{}$possession|@douglas2004natural|yes, see above|
|monotheism|@wood2018complexity|absolutely, the variable is right there|

At this point, we can say, okay, the model says that these various societal covariates are meaningful in the Atlas. But the direction of the individual relationships is not clear. If our analysis finds that increased prevalence of slavery makes possession trance less likely, it's clearly at odds with the existing literature and probably doesn't make sense.

## 7. Making sense: a figure

The figure below shows the proportion of societies with possession trance (dark blue) and without (light blue) across the ordered levels of the predictors that the model thinks are important. They come in order of how important they are, according to the model; prevalence of slavery, then family organisation, and so on. The predictor levels are ordered according to coding and generally go from less prevalent (top) to more prevalent (bottom):

```{r fig.width=10,fig.height=15}
d_out = d %>% 
  filter(var_id == 'EA112') %>% 
  mutate(
    possession_trance_present = case_when(
      code %in% c(3,4,6,7) ~ 'present',
      code %in% c(1,2,5,8) ~ 'absent'
        )
      ) %>% 
  select(soc_id,possession_trance_present)

d_in = d %>%
  filter(var_title %in% vars2$Name, !is.na(code)) %>% 
  mutate(
    predictor = fct_reorder(var_name, -as.double(recoded_value)),
    var_title = fct_relevel(var_title, vars2$Name)
    ) %>% 
  select(soc_id,var_definition,var_title,predictor)

d_vis = inner_join(d_in,d_out, by = 'soc_id')

plots = d_vis %>%
  count(var_title,predictor,possession_trance_present) %>% 
  # pivot_wider(names_from = possession_trance_present, values_from = n, values_fill = 0) %>% 
  # mutate(log_odds_pt = log((`TRUE` + 1) / (`FALSE` + 1)) ) %>% 
  group_by(var_title) %>% 
  nest() %>% 
  mutate(
    plot = map2(var_title, data, ~ ggplot(.y, aes(predictor,n,fill = possession_trance_present)) +
                  geom_col(position = 'dodge') +
                  theme_bw() +
                  theme(axis.title=element_blank()) +
                  guides(fill = 'none') +
                  scale_fill_brewer(palette = "Paired") +
                  ggtitle(str_wrap(.x, width = 25)) +
                  coord_flip()
                 )
  ) %>% 
  pull(plot)

wrap_plots(plots, ncol = 2) + plot_annotation(caption = 'Dark blue: possession trance present. Light blue: absent.')
```

For the most important variables, we see a clear correllation between the presence of possession trance and the variable scaling up. Societies in which slavery is more prevalent have more possession trance than those where it is less prevalent. Same is true for societies with more complex family structures, more social complexity, more complex settlement patterns, and larger settlements.

The societal variables that do not, strictly speaking, measure social organisation are less obvious. Apparently cousin marriage is good for possession trance, and so is having a lot of words for your various cousins. (We might suspect the latter is here because it is heavily correlated with the former.) Active, moral high gods are bad for possession trance. Prohibitions on pre-married sexual behaviour for women go with more prevalent possession trance, but this pattern is rather variable.

## 8. Summary

We can predict some types of trance and possession phenomena in the SCCS subset of the EA. Not all of them. Some of the important co-variates are the ones mentioned by earlier work and the direction of the relationships corroborates the literature.

Human society is complicated and cross-cultural data have very low resolution. In this sense, we'd be very suspect of a model that would be too accurate of predicting possession trance from societal covariates. It cannot replace fieldwork and more complex explanatory models either. What it can do is provide an extra leg for such explanatory work, which is broadly the recognised ambition of cross-cultural comparative work in evolutionary anthropology.

Our approach is statistically strict, so the fact that it still shows some accuracy is definitely promising for the validity of the connections drawn by previous work on trance and possession phenomena in particular and large cross-cultural comparisons in general.

## References

### 2022-06-24 helper.R

@@ -1,0 +1,248 @@
+# pulls data in from the DPLACE github repository, filters for trance & possesion analysis, tidies it up, saves it
+# author: pracz
+# date: 22/2/22
+
+setwd('~/Github/Racz2023trance/')
+library(tidyverse)
+library(glue)
+library(caret)
+library(rjson)
+library(RCurl)
+library(vctrs)
+
+# --- pull --- #
+
+# pull variables codes etc from the dplace github repository
+variables = read_csv("https://raw.githubusercontent.com/D-PLACE/dplace-data/master/datasets/EA/variables.csv")
+
+codes = read_csv("https://raw.githubusercontent.com/D-PLACE/dplace-data/master/datasets/EA/codes.csv")
+
+codes = codes %>% # we need some changes in codes
+  rename(
+    var_description = description,
+    var_name = name
+  ) %>% 
+  mutate(
+    num_id = str_replace(var_id, 'EA', '') %>% as.double() %>% as.character()
+  )
+
+dat = read_csv("https://raw.githubusercontent.com/D-PLACE/dplace-data/master/datasets/EA/data.csv")
+
+languages = read_csv("https://raw.githubusercontent.com/D-PLACE/dplace-data/master/csv/glottolog.csv") %>% 
+  rename(family = family_name, glottocode = id) %>% 
+  select(family, glottocode)
+
+sccs = read_csv("https://raw.githubusercontent.com/D-PLACE/dplace-data/master/datasets/SCCS/societies.csv") %>% 
+  select(pref_name_for_society,glottocode,HRAF_name_ID) %>% 
+  rename(society = pref_name_for_society)
+
+# for societies, we don't keep everything and rename some things
+societies = read_csv("https://raw.githubusercontent.com/D-PLACE/dplace-data/master/datasets/EA/societies.csv") %>% 
+  select(id, glottocode, pref_name_for_society, Lat, Long) %>% 
+  rename(
+    lat = Lat, 
+    lon = Long, 
+    soc_id = id, 
+    society = pref_name_for_society
+  )
+
+# locations is a json. the json converter doesn't quite cope with it so we need some tricks
+locations = getURL("https://raw.githubusercontent.com/D-PLACE/dplace-data/master/geo/societies_tdwg.json") %>%
+  fromJSON # makes it into a list of lists
+soc_id = names(locations)
+locations = locations %>% 
[…]

### 2022-06-24 model_gbm.R

@@ -1,0 +1,315 @@
+################################################
+# gbm
+################################################
+
+set.seed(1337)
+
+# -- header -- #
+
+setwd('~/Github/Racz2023trance/')
+
+library(tidyverse)
+library(magrittr)
+library(glue)
+library(h2o) # to use h2o
+# don't panic:
+# https://docs.h2o.ai/h2o/latest-stable/h2o-docs/index.html
+
+# -- read-in -- #
+
+d = read_tsv('data/dat_wide.tsv')
+
+# -- wrangling -- #
+
+# outcome = d %>% 
+#   filter(var_id == 'EA112') %>% 
+#   mutate(
+#     outcome = case_when(
+#       var_description == 'No trance states of any kind are known to occur, and there is no belief in possession.' ~ 'No trance states nor belief in possession',
+#       var_description == 'Trance behavior is known to occur, but there is no belief in possession.' ~ 'Trance only',
+#       var_description == 'A belief in possession exists.' ~ 'Possession only',
+#       var_description == 'There is both a trance state and a belief in possession, but this belief refers to phenomena other than trance, which is explained through other categories.' ~ 'Separate trance and possession',
+#       var_description %in% c(
+#         'Two types of trance states are known to occur. One which is explained as due to possession and one which is given another type of explanation. In addition to explaining trance, possession belief also refers to one or more other phenomena.',
+#         'Trance explained as due to possession is known to occur, and there are no other trance states, but cases of possession outside of trance are also believed to occur.',
+#         'Trance states of two kinds are known to occur, some of which are explained by possession. No other phenomena are explained by possession.'
+#         ) ~ 'Trance typically if possession',
+#       var_description == 'Trance behavior is known to occur and is explained as due to possession. There is no possession belief referring to other experiences and there are no trance states with other explanations.' ~ 'Trance if and only if possession'
+#     )
+#   ) %>% 
+#   select(soc_id,outcome)
+
[…]

### 2022-10-20 Readme.Rmd
reran booster

@@
 title: Identifying covariate traits of trance and possession phenomena in the Ethnographic
-  Atlas using sparse regression
+  Atlas using boosting and nested regression
 author: "Péter Rácz"
@@
 
-d = read_tsv('data/dat_long.tsv')
-sccs = filter(d, in_sccs)
+ld = read_tsv('data/dat_long.tsv')
+# wd = read_tsv('data/dat_wide.tsv')
+sccs = filter(ld, in_sccs)
 ```
@@
 We want to look at two questions.
-What sort of social factors have been proposed to cooccur with trance and possession phenomena?
-Using the Ethnographic Atlas and the Standard Cross-Cultural Sample, can we get more confident that these factors do cooccur with such phenomena across cultures?
+What sort of social factors have been proposed to co-occur with trance and possession phenomena?
+Using the Ethnographic Atlas and the Standard Cross-Cultural Sample, can we get more confident that these factors do co-occur with such phenomena across cultures?
 
@@
 
-Many people have argued that trance induced by possession is not a random, idiosynchratic pathology but a recurrent psychosocial trait that is at least partly motivated by other social factors.
+Many people have argued that trance induced by possession is not a random, idiosyncratic pathology but a recurrent psycho-social trait that is at least partly motivated by other social factors.
 
@@
 
-Several works made the same point or emphasised various aspects of the same relationship. The above table is a bit reductive.
-
-It's an entirely different question whether we can build more ambitious explanatory models that link possession trance and similar practices to e.g.
-social inequality.
-This is not a question we really want to deal with here.
-What we want to know is whether the above descriptions are accurate at all, when given closer scrutiny.
-We can answer this question by looking at a cross-cultural data set, the Ethnographic Atlas, to see if we can attest in it any of the correlations proposed in the literature.
+Several works made the same point or emphasised various aspects of the same relationship. The above table is somewhat reductive.
 
@@
 
-We find data on trance and possession phenomena and their various proposed covariates in the Ethnographic Atlas [@murdock1967ethnographic] (we use the version made available in the D-Place database, @kirby2016d).
+The use of cross-cultural typological data in general and the Ethnographic Atlas in particular has precedents.
+@bourguignon1968, @bourguignon1973religion, and @greenbaum1973 use the Ethnographic Atlas to look for correlates of trance and possession trance.
+@wood2018complexity, @shaara1992, @winkelman1986, and @douglas2004natural build on this research by specifying and compiling their own data sets.
+The present paper uses the Ethnographic Atlas as made available in the D-Place database, @kirby2016d.
 
@@
 ```{r }
-d %>% 
+ld %>% 
   filter(var_title == 'Trance states', !is.na(code)) %>% 
@@
 
-The eight levels here likely result from a trade-off between description accuracy and generalisability that is present throughout the Atlas.
-It is unlikely that eight descriptions suffice to capture the plethora of phenomena related to trance and possession.
-But largely the same is true for all variables in the Atlas.
-This hasn't stopped people from doing large-scale comparative studies before!
-
-The Atlas lists a set of variables that tie in with the observed covariates in the literature. Here are some relevant-looking ones:
+The Atlas lists a set of variables that tie in with the observed covariates in the literature.
+These are either taken literally from @bourguignon1968, @bourguignon1973religion, and @greenbaum1973 or adapt variables used by @wood2018complexity, @shaara1992, @winkelman1986, and @douglas2004natural.
 
 ```{r }
-d %>% 
+ld %>% 
   filter(var_title != 'Trance states') %>% 
@@
 
-The covariates in the literature do not directly map on the variables in the Atlas.
-The variables are themselves correlated, either because they measure different facets of the same thing (like how complex a society is) or because they were adapted together by socially or geographically close societies (Galton's problem).
-
 ## 3. How do we look for correlations between trance and possession and their proposed societal covariates?
 
-We could table how often trance and possession phenomena and their posited covariates pattern together. This is generally a bad idea.
-
-There are eight types of trance and possession phenomena recorded in the Atlas (and we can make up new codings!) and a wide set of possible covariates, all with a number of levels:
+There are eight types of trance and possession phenomena recorded in the Atlas and a wide set of possible covariates, all with a number of levels.
 
 ```{r }
-vars = d %>% 
+vars = ld %>% 
   filter(var_title != 'Trance states') %>% 
@@
 
-Even if we initially restrict ourselves by a hypothesis, if we keep trying all the possible combinations, we are bound to find something.
+The data are challenging in two ways.
+The variables are themselves correlated, either because they measure different facets of the same thing (like how complex a society is) or because they were adapted together by socially or geographically close societies (Galton's problem).
 In addition, data are missing from the Atlas in a non-random manner (e.g.
 if a society is hard to access, it is harder to provide a population size estimate for it).
-This makes any catch even more spurious.
 
-We could assess the probability of all eight descriptions in the Atlas together by using a multinomial model.
-But because data are missing for the covariates and because the covariates are themselves correlated, a single model will be wasteful (losing all the observations for which data are missing) and badly fit (with a lot of colinear predictors in it).
-We're probably better off building a lot of models and then aggregating over them to find the most important covariates of trance and possession phenomena.
+We can account for Galton's problem (geographic and cultural co-variation) by restricting our initial data to societies in the Standard Cross-Cultural Sample.
 
-We can do this by using an ensemble learning method.
-This means that we start with societies with information on trance and possession phenomena.
-We build a lot of tiny models, using random samples of the data and random subsets of our covariates.
-Then, we compare these to assess the overall importance of our covariates.
+The remaining co-variation and the non-random patterns of missing data recommends an ensemble learning method.
+This starts with societies with information on trance and possession phenomena and builds a lot of tiny models, using random samples of the data and random subsets of the covariates.
+Then, we compare these models to assess the overall importance of our covariates.
 Building little models in parallel is called bagging.
 This is how a random forest works, growing many classification or regression trees and aggregating over the results.
-One step further is to train little models on the mistakes of the previous little models.
+One step further is to train models on the mistakes of the previous models.
 This is called [boosting](https://www.ibm.com/cloud/learn/boosting) and usually results in higher training and prediction accuracy.
@@
 
-We only fit the model on societies which (a) have information on whether trance and possession phenomena are present (we can't do much with those where we don't know this) and (b) come from the standard cross-cultural sample, which minimises the phylogenetic signal so we don't have to worry about it.
+We only fit the model on societies which (a) have information on whether trance and possession phenomena are present and (b) come from the standard cross-cultural sample.
+
+Following @bourguignon1968, we reduce the complexity of the outcome variable to two levels: whether a society has possession trance or not.
+The predictors are the possible covariates identified above.
+Some predictors, like `EA032` (Jurisdictional hierarchy of local community) or `EA113` can be construed to follow a scale.
+These are input as ordered factors.
+Others, like `EA053` (Sex differences in animal husbandry) or `EA043` (Major type of descent) can be carved up into categories: These are one-hot encoded.
+For details, see the helper file.
 
 We use hyperparameter tuning to find the best possible model.
-The outcome of the model is one of the eight descriptions of trance and possession phenomena in the Atlas.
-The predictors are the possible covariates identified above.
 The model is fit on a training set of `length(unique(sccs$soc_id))` societies with information on trance and possession in the standard cross-cultural sample.
-
-I'm scarce on the technical details here on purpose, but the analysis can be replicated with the code in this repository.
 

### 2022-10-20 helper.R

@@ -1,6 +1,8 @@
 # pulls data in from the DPLACE github repository, filters for trance & possesion analysis, tidies it up, saves it
 # author: pracz
-# date: 22/2/22
+# date: 20/10/22
+
+set.seed(1337)
 
 setwd('~/Github/Racz2023trance/')
 library(tidyverse)
@@ -73,7 +75,7 @@ dat = dat %>%
 # --- filter --- #
 
 # these are the variables of interest for trance and possession
-keep_vars = paste('EA', c("034","012","008","015","023","043","027","031","030","032","033","042","066","068","070","072","074","078","113","112"), sep='')
+keep_vars = paste('EA', c("008","012","015","023","027","030","031","032","033","034","042","043","053","054","066","068","070","072","073","074","078","112","113"), sep='')
 
 # we recode some of these as ordinal variables, others as factors, we leave the rest as is, thereby assuming it is an ordinal variable
 trance_dict = codes %>% 
@@ -103,10 +105,22 @@ trance_dict = codes %>%
         var_id == 'EA042' & code %in% c(4) ~ "pastoralism",
         var_id == 'EA042' & code %in% c(1,2,3) ~ "foraging",
         var_id == 'EA042' & code == 8 ~ 'multiple',
+        var_id == 'EA053' & code %in% c(1,2) ~ 'mostly male',
+        var_id == 'EA053' & code %in% c(5,6) ~ 'mostly female',
+        var_id == 'EA053' & code %in% c(3,4) ~ 'both',
+        var_id == 'EA053' & code %in% c(7,8,9) ~ 'other',
+        var_id == 'EA054' & code %in% c(1,2) ~ 'mostly male',
+        var_id == 'EA054' & code %in% c(5,6) ~ 'mostly female',
+        var_id == 'EA054' & code %in% c(3,4) ~ 'both',
+        var_id == 'EA054' & code %in% c(7,8,9) ~ 'other',
         var_id == 'EA072' & code %in% c(1) ~ 'patrilineal',
         var_id == 'EA072' & code %in% c(2) ~ 'matrilineal',
         var_id == 'EA072' & code %in% c(3,4,5,6,7) ~ 'other',
         var_id == 'EA072' & code %in% c(9) ~ 'absent',
+        var_id == 'EA073' & code %in% c(1,2) ~ 'patrilineal',
+        var_id == 'EA073' & code %in% c(3,4) ~ 'matrilineal',
[…]

### 2022-10-20 model_gbm.R

@@ -44,18 +44,14 @@ d %<>%
   mutate_if(is.character, as.factor) %>% 
   mutate_if(is.logical, as.double)
 
+d$possession_trance_present = as.factor(d$possession_trance_present)
+
 d_sccs = filter(d, in_sccs == 1)
 
 x = names(d)[!names(d) %in% c('soc_id','outcome','EA112_Trance_states','trance_present','possession_present','possession_trance_present','in_sccs')]
-y = 'outcome'
+y = 'possession_trance_present'
 
 # -- gbm -- #
-
-# four gbm models. 
-# i everything
-# ii trance only
-# iii possession only
-# iv possession trance only
 
 h2o.init(nthreads=16)
 
@@ -71,10 +67,6 @@ gbm_params1 = list(
   min_split_improvement = c(1e-3,1e-5)
 )
 
-#
-# i all possible levels
-#
-
 # we do a grid search
 gbm_grid1 = h2o.grid("gbm", 
                      x = x, 
@@ -84,232 +76,52 @@ gbm_grid1 = h2o.grid("gbm",
                      nfolds = 3,
                      ntrees = 100,
                      seed = 1,
-                     distribution = 'multinomial',
+                     distribution = 'bernoulli',
                      categorical_encoding = 'enum', # important
                      stopping_metric = 'logloss',
                      hyper_params = gbm_params1
 )
 
-# save grid
-# h2o.saveGrid(
-#   'models',
-#   'gbm_grid1',
-#   save_params_references = FALSE,
-#   export_cross_validation_predictions = FALSE
-# )
-# gbm_grid = h2o.loadGrid('models/gbm_grid')
+h2o.saveGrid(grid_directory = 'grid', grid_id = 'gbm_grid1')
 
 # pick the best models from grid
 gbm_grid_best1 = h2o.getGrid(grid_id = "gbm_grid1",
-                            sort_by = "logloss",
-                            decreasing = FALSE)
-
-# pick bestest model
-fit11 = h2o.getModel(gbm_grid_best1@model_ids[[1]])
-# pick worst model
-fit12 = h2o.getModel(gbm_grid_best1@model_ids[[384]])
-
-# check stats
-h2o.performance(fit11, xval = TRUE)
-h2o.performance(fit12, xval = TRUE)
-
-# varimp
-varimp11 = as_tibble(h2o.varimp(fit11))
-
-# don't trust the built-in confusion matrix!
[…]

### 2022-10-20 Readme.Rmd
revamped analysis

@@
 
-Here is the distribution of the various descriptions for trance and possession in our training sample. I gave the descriptions shorter names as well, to make things easier later on:
+Here is the distribution of the various descriptions for trance and possession in our training sample.
 
 ```{r }
-sccs %>% 
+trance = sccs %>% 
   filter(var_id == 'EA112') %>% 
-  count(var_name,var_description) %>% 
-  rename(name = var_name, description = var_description, count = n) %>% 
+  count(var_name,code,var_description) %>% 
+  rename(name = var_name, description = var_description, count = n)
+
+trance %>% 
+  select(-code) %>% 
   add_row(name = 'Total', description = '', count = 151) %>% 
@@
 
-We see that only about 10% of societies have no trance or possession phenomena. From the rest, a majority have possession trance in some form.
+We see that only about 10% of societies have no trance or possession phenomena.
+From the rest, a majority have possession trance in some form.
+Following @bourguignon1968 we can reduce this to a binary: 81 societies have possession trance, 70 have other forms of trance and possessior or show neither.
 
-We compare this with how our model categorises each society.
+```{r }
+trance %>% 
+  mutate(
+    possession_trance = case_when(
+      code %in% c(3,4,6,7) ~ 'possession trance',
+      code %in% c(1,2,8,5) ~ 'other'
+    ) %>% 
+      fct_relevel('possession trance')
+  ) %>% 
+  group_by(possession_trance) %>% 
+  summarise(count = sum(count)) %>% 
+  kable('simple')
+```
+
+We compare this with how our best model categorises each society.
 This is our model's confusion matrix.
-The rows are the actual descriptions, the columns are the predicted descriptions.
-So a good model would have large counts on the diagonal (where it got the label right) and low values elsewhere (where it didn't).
-The last two columns give the model's error as a probability (0-1) and as a rate (as n/m).
+`0` is absence of possession trance, `1` is presence of possession trance.
+The rows add up to values in the original, the columns are the model's predictions.
+The last two columns show the error rate.
+The model's accuracy on absence of possession trance is relatively low, though higher than chance (54 out of 80, p < 0.01 using a binomial test with a true p of .529).
+It is mostly accurate capturing the presence of possession trance.
 
@@
   kable('simple', digits = 2)
+
+# precision = true 1/false 1
+# recall = true 1/true1 + false0
+precision = 68 / 94
+recall = 68 / 71
+f = (2 * precision * recall) / (precision + recall)
 ```
 
-These results are not very impressive.
-The model gets some of the larger categories less wrong but it is completely off the mark on the smaller ones.
-(It gets only one society right in the "no trance or possession" category and all of them wrong in the "possession-only" category!)
+The model has an F-measure of `r round(f,2)`, with a penalty on overgeneralising possession trance.
+Though comparisons are hard to make, this is not particularly accurate for a binary categorisation problem.
+We did expect this, though, as it is unlikely that any complex cultural phenomenon could be reduced to a small number of cross-cultural predictors.
 
-Focussing on the relationship between category sizes and model accuracy, we can tell two stories here.
-The first one is that the model is terrible and it is only likely to do better with the larger categories because of their size (so that the baseline probability of category membership is higher).
-The second story is largely the same but with some hopeful notes.
+### 6. Results on the EA data
 
-We can say that the way trance and possession phenomena are defined in the Atlas is very fragmentary.
-It involves at least two external sources of noise.
-First, for a category like "trance only, no possession", maybe the curators of the Atlas put very disparate psychosocial phenomena into this category, which haven't a lot to do with each other at all.
-In this case, it is no small wonder that we can't capture this category algorithmically.
-Maybe it's all arbitrary!
+We can ask the model to make predictions for the entire Ethnographic Atlas (where trance and possession data are available) -- including the SCCS subset.
+Generally, you want to separate the training and test sets in Machine Learning and not do this, but the 'everything except the SCCS' subset of the Atlas would be an odd test set.
 
-Second, maybe, even for the larger categories, the boundaries are porous.
-Maybe it's not entirely clear whether, in a given society, only possession explains trance phenomena.
-Maybe the fieldworkers missed something!
-Maybe they disagree on their definitions of things (e.g.
-some say that a trance phenomenon in a given society is not caused by possession ergo it is not trance at all but something else).
-
-We can work around some of these issues by re-defining our outcome variable.
-We split our descriptions into two categories, based on whether they capture possession trance or not.
-This would look like this:
+Here is the confusion matrix for all `r length(unique(ld$soc_id))` societies in the Atlas where trance and possession information are available:
 
 ```{r }
-d %>% 
-  filter(var_title == 'Trance states') %>% 
-  distinct(var_description) %>% 
-  rename(description = var_description) %>% 
-  add_column(`this is...` = c('something else','something else','possession trance','possession trance','possession trance','something else','something else','possession trance')) %>% 
+cm1b = read_tsv('models/conf_matrix11test.tsv')
+
+cm1b %>% 
   kable('simple')
+
+# precision = true 1/false 1
+# recall = true 1/true1 + false0
+precision2 = 221 / 389
+recall2 = 221 / 281
+f2 = (2 * precision2 * recall2) / (precision2 + recall2)
+
 ```
 
-Now we have two categories that are roughly the same size and are perhaps less conceptually ambiguous.
-We refit the model, but this time it only tries to guess whether a society has possession trance or something else (including no trance or possession at all).
+The F-measue of this model is `r round(f2,2)`.
+Accuracy here is much worse than for the training data only.
+That is at least partly because the model is ignorant of the phylogenetic signal which might result in patterns that are unexpected from a purely correlational point of view.
+However, the model is still more accurate than it would be by chance.
 
-## 6. Results on the training sample, part II
-
-We look at the model's confusion matrix:
+We take a look at the variables that were most important in the model in predicting the presence of possession trance in a given society, with a cutoff of 10/47.
 
 ```{r }
-cm4a = read_tsv('models/conf_matrix41model.tsv')
-cm4a %>% 
-  mutate(original = c(names(cm4a)[!names(cm4a) %in% c('Error','Rate')],'Total')) %>% 
-  relocate(original) %>% 
-  kable('simple', digits = 2)
-```
-
-This is a lot better.
-Granted, there are only two categories, but the error rate is much lower.
-This means our model can more or less use the societal covariants to guess whether a society, by the Atlas' own terms, will have possession trance phenomena (1) or not (0).
-
-Having established this, we can ask the model to make predictions for the entire Ethnographic Atlas (where trance and possession data are available, of course) -- including the SCCS subset.
-Generally, you want to separate the training and test sets in Machine Learning and not do this, but the 'everything except the SCCS' subset of the Atlas would be an odd test set.
-
-Here is the confusion matrix for all `r length(unique(d$soc_id))` societies in the Atlas where trance and possession information are available:
-
-```{r }
-cm4b = read_tsv('models/conf_matrix41test.tsv')
-
-cm4b %>% 
-  add_column(original = c(0,0,0,1,1,0,1,1)) %>% 
-  group_by(original) %>% 
-  summarise(`0` = sum(`0`), `1` = sum(`1`)) %>% 
-  kable('simple')
-
-# d %>% 
-#   filter(var_id == 'EA112') %>% 
-#   mutate(
-#     possession_trance_present = case_when(
-#       code %in% c(3,4,6,7) ~ T,
-#       code %in% c(1,2,5,8) ~ F
-#         )
-#   ) %>% 
-#   count(possession_trance_present) %>% 
-#   kable('simple')
-```
-
-This is much worse than the training data only but that is at least partly because the model is ignorant of the phylogenetic signal which might result in patterns that are unexpected from a purely correlational point of view.
-And it is still not terrible!
-It gets 468/`r length(unique(d$soc_id))` societies right.
-There are 281 versus 377 societies in the data with and without possession trance.
-If it assumed that no societies had possession trance, it would get 377 data right, and 468 is considerably larger number than that.
-This means the model is doing something right.
-
-Now that we established that the model does find some purchase, we take a look at the variables that were most important in predicting the presence of possession trance in a given society.
-I will only show societal covariants with a high enough importance:
-
-```{r }
-varimp4 = read_tsv('models/varimp41.tsv')
-vars2 = varimp4 %>% 
-  filter(relative_importance > 15) %>%
+varimp1 = read_tsv('models/varimp11.tsv')
+vars2 = varimp1 %>% 
+  slice(1:10) %>% 
   mutate(ID = str_extract(variable, '^EA...')) %>% 
@@
 
-What do we see?
-The five most important driving factors are the prevalence of slavery, family complexity, the complexity of class structure, premarital norms for women, and settlement patterns.
-Various aspects of social complexity and filial and consanguineal organisation also make the list.
-We can cross-reference this with our first table, with pointers in the literature.
-We add a column indicating whether the Atlas seems to support each specific claim.
-(This we do in a very haphazard manner!)
-
-| Relationship|Reference|Our model|
-|:-|:-|:-|
-|less complexity$\rightarrow{}$trance w/o possession|@bourguignon1968|not directly, but more complexity means possession trance is more likely|
-|women's social participation|@bourguignon1973religion|not directly, but premarital sexual norms can be a proxy|
-|increased role differentiation and slavery practices$\rightarrow$trance and possession|@greenbaum1973|slavery, social and class complexity are relevant|
-|main mode of subsistence|@bourgouignon1977|not really|
-|increased social participation$\rightarrow{}$possession|@swanson1978|indirectly, through domestic and social complexity|
-|more complexity$\rightarrow{}$trance w/ possession|@winkelman1986|yes, see above|
-|cognatic kinship$\rightarrow{}$more possession and trance|@shaara1992|not directly, but family and kinship seems to do something|
-|decreased social participation$\rightarrow{}$possession|@douglas2004natural|yes, see above|
-|monotheism|@wood2018complexity|absolutely, the variable is right there|
-
-At this point, we can say, okay, the model says that these various societal covariates are meaningful in the Atlas.
-But the direction of the individual relationships is not clear.
-If our analysis finds that increased prevalence of slavery makes possession trance less likely, it's clearly at odds with the existing literature and probably doesn't make sense.
-
-## 7. Making sense: a figure
+We now look at the direction of these effects (does e.g.\ increased jurisdictional hierarchy go with more or less prevalent possession trance?).
 
@@
 
+For the most important variables, we see a clear correllation between the presence of possession trance and the variable scaling up.
+Societies in which slavery is more prevalent have more possession trance than those where it is less prevalent.
+Same is true for societies with more complex family structures, more social complexity, more complex settlement patterns, and larger settlements.
+The societal variables that do not, strictly speaking, measure social organisation are less obvious.
+
 ```{r fig.width=10,fig.height=15}
-d_out = d %>% 
+d_out = ld %>% 
   filter(var_id == 'EA112') %>% 
@@
 
-d_in = d %>%
+d_in = ld %>%
   filter(var_title %in% vars2$Name, !is.na(code)) %>% 
@@
 
-For the most important variables, we see a clear correllation between the presence of possession trance and the variable scaling up.
-Societies in which slavery is more prevalent have more possession trance than those where it is less prevalent.
-Same is true for societies with more complex family structures, more social complexity, more complex settlement patterns, and larger settlements.
+## 7. Summary
 
-The societal variables that do not, strictly speaking, measure social organisation are less obvious.
-Apparently cousin marriage is good for possession trance, and so is having a lot of words for your various cousins.
-(We might suspect the latter is here because it is heavily correlated with the former.) Active, moral high gods are bad for possession trance.
-Prohibitions on pre-married sexual behaviour for women go with more prevalent possession trance, but this pattern is rather variable.
+These results go some way in confirming hypotheses on the cross-cultural correlates of possession trance in the literature.
+@bourguignon1968 explicitly tests the variables of slavery, jurisdictional hierarchy, settlement patterns, and kin groups.
+@shaara1992 point to a high value placed on female virginity at marriage, captured here by norms of premarital sexual behaviour for girls.
+These are all important variables in our model.
+@greenbaum1973 and @shaara1992 point to sexual division of labour, while @wood2018complexity to the role of moralising high gods, neither of which is relevant in our model.
 
-## 8. Summary
+This brings us back to the original predictions in the table above.
+
+| Relationship|Reference|Our model|
+|:-|:-|:-|
+|less complexity$\rightarrow{}$trance w/o possession|@bourguignon1968|the same measures of complexity are relevant|
+|women's social participation|@bourguignon1973religion|labour divisions are not relevant|
+|increased role differentiation and slavery practices$\rightarrow$trance and possession|@greenbaum1973|slavery, social and class complexity are relevant|
+|main mode of subsistence|@bourgouignon1977|not relevant|
+|increased social participation$\rightarrow{}$possession|@swanson1978|indirectly, through domestic and social complexity|
+|more complexity$\rightarrow{}$trance w/ possession|@winkelman1986|yes, see above|
+|cognatic kinship$\rightarrow{}$more possession and trance|@shaara1992|not directly, but kinship structure is relevant|
+|decreased social participation$\rightarrow{}$possession|@douglas2004natural|yes, see above|
+|monotheism|@wood2018complexity|not relevant in our model|
 

### 2022-10-20 Readme.Rmd
some more work

@@
 
-There are eight types of trance and possession phenomena recorded in the Atlas and a wide set of possible covariates, all with a number of levels.
-
-```{r }
-vars = ld %>% 
-  filter(var_title != 'Trance states') %>% 
-  distinct(var_id,var_title,var_name) %>%
-  arrange(var_id) %>% 
-  group_by(var_id, var_title) %>% 
-  nest() %>% 
-  mutate(levels = map(data, ~ paste(.$var_name, collapse = ', '))) %>% 
-  select(var_id, var_title, levels) %>% 
-  unnest(cols = c(levels)) %>% 
-  rename(ID = var_id, Name = var_title)
-
-kable(vars, 'simple')
-```
+There are eight types of trance and possession phenomena recorded in the Atlas and a wide set of possible covariates, all with a number of levels (see the Appendix).
 
@@
 Others, like `EA053` (Sex differences in animal husbandry) or `EA043` (Major type of descent) can be carved up into categories: These are one-hot encoded.
-For details, see the helper file.
+For details, see the Appendix.
 
@@
 ```{r }
+vars = ld %>% 
+  filter(var_title != 'Trance states') %>% 
+  distinct(var_id,var_title,var_definition,var_name) %>%
+  arrange(var_id) %>% 
+  group_by(var_id, var_title, var_definition) %>% 
+  nest() %>% 
+  mutate(levels = map(data, ~ paste(.$var_name, collapse = ', '))) %>% 
+  select(var_id, var_title, levels) %>% 
+  unnest(cols = c(levels)) %>% 
+  rename(ID = var_id, Name = var_title)
+
 varimp1 = read_tsv('models/varimp11.tsv')
@@
 
+## Appendix
+
+### Relevant predictors in the Ethnographic Atlas
+
+A survey of the literature resulted in the following predictors of trance and possession phenomena.
+
+```{r }
+kable(vars, 'simple')
+```
+
+### Recoded values in the Ethnographic Atlas
+
+The recoded values were used in the model for each predictor.
+Where the recoded values are numeric, the predictor was used as an ordered factor.
+Where the recoded values are strings, the predictor was one-hot encoded (such that predictor A with levels 1, 2, and 3 was turned into boolean predictors A==2 and A==3).
+
+```{r }
+ld %>% 
+  filter(var_id != 'EA112') %>% 
+  distinct(var_id,var_title,var_description,recoded_value) %>% 
+  arrange(var_id,recoded_value) %>% 
+  kable('simple')
+```
+
 ## References

### 2022-10-20 Readme.Rmd
some edits

@@
   count(var_name,code,var_description) %>% 
-  rename(name = var_name, description = var_description, count = n)
+  rename(name = var_name, description = var_description, count = n) %>% 
+  mutate(
+    possession_trance = case_when(
+      code %in% c(3,4,6,7) ~ 'possession trance',
+      code %in% c(1,2,8,5) ~ 'other'
+    ) %>% 
+      fct_relevel('possession trance')
+  )
 
 trance %>% 
-  select(-code) %>% 
+  select(description,code,name,possession_trance,count) %>%
+  rename(outcome = possession_trance) %>% 
   add_row(name = 'Total', description = '', count = 151) %>% 
@@
 trance %>% 
-  mutate(
-    possession_trance = case_when(
-      code %in% c(3,4,6,7) ~ 'possession trance',
-      code %in% c(1,2,8,5) ~ 'other'
-    ) %>% 
-      fct_relevel('possession trance')
-  ) %>% 
   group_by(possession_trance) %>% 

### 2022-10-21 Readme.Rmd
working on paper in gdocs

--- 2022-10-20 Readme.Rmd
@@
 ```{r }
+trance = sccs %>% 
+  filter(var_id == 'EA112') %>% 
+  count(var_name,code,var_description) %>% 
+  rename(name = var_name, description = var_description, "n in SCCS" = n) %>% 
+  mutate(
+    possession_trance = case_when(
+      code %in% c(3,4,6,7) ~ 'possession trance',
+      code %in% c(1,2,8,5) ~ 'other'
+    ) %>% 
+      fct_relevel('possession trance')
+  )
+
 ld %>% 
   filter(var_title == 'Trance states', !is.na(code)) %>% 
-  distinct(code, var_description) %>% 
+  count(code, var_description, name = 'n in EA') %>% 
+  left_join(trance) %>% 
   arrange(code) %>% 
-  rename(coding = code, Description = var_description) %>% 
+  rename(Coding = code, Description = var_description) %>% 
+  select(Coding,Description,possession_trance,`n in EA`,`n in SCCS`) %>% 
   kable('simple')
@@
 
-Here is the distribution of the various descriptions for trance and possession in our training sample.
-
-```{r }
-trance = sccs %>% 
-  filter(var_id == 'EA112') %>% 
-  count(var_name,code,var_description) %>% 
-  rename(name = var_name, description = var_description, count = n)
-
-trance %>% 
-  select(-code) %>% 
-  add_row(name = 'Total', description = '', count = 151) %>% 
-  kable('simple')
-```
-
-We see that only about 10% of societies have no trance or possession phenomena.
+Looking back at the table in the previous section, we see that only about 10% of societies have no trance or possession phenomena.
 From the rest, a majority have possession trance in some form.
@@
 trance %>% 
-  mutate(
-    possession_trance = case_when(
-      code %in% c(3,4,6,7) ~ 'possession trance',
-      code %in% c(1,2,8,5) ~ 'other'
-    ) %>% 
-      fct_relevel('possession trance')
-  ) %>% 
   group_by(possession_trance) %>% 
-  summarise(count = sum(count)) %>% 
+  summarise(count = sum(`n in SCCS`)) %>% 
   kable('simple')

### 2022-10-21 Readme.Rmd
whoopsie daisy

@@
 ---
+
+# To do: one-hot encoding ended up creating unidentifiable vars (where both values of a col are present) so do something with that.
 
@@
 ld = read_tsv('data/dat_long.tsv')
-# wd = read_tsv('data/dat_wide.tsv')
+wd = read_tsv('data/dat_wide.tsv')
 sccs = filter(ld, in_sccs)
@@
 |:-----------------------------------|:-----------------------------------|
-| less complexity $\rightarrow{}$ trance w/o possession                                    | @bourguignon1968         |
-| women's social participation                                                             | @bourguignon1973religion |
-| increased role differentiation and slavery parctices $\rightarrow$ trance and possession | @greenbaum1973           |
-| main mode of subsistence                                                                 | @bourgouignon1977        |
-| increased social participation $\rightarrow{}$ possession                                | @swanson1978             |
-| more complexity $\rightarrow{}$ trance w/ possession                                     | @winkelman1986           |
-| cognatic kinship $\rightarrow{}$ more possession and trance                              | @shaara1992              |
-| decreased social participation $\rightarrow{}$ possession                                | @douglas2004natural      |
-| monotheism                                                                               | @wood2018complexity      |
+| Social complexity, rigidity, gender role differentiation                                    | @bourguignon1968         |
+| Women's social participation                                                             | @bourguignon1973religion |
+| Slavery, role differentiation, and structure differentiation | @greenbaum1973           |
+| Increased social participation                                | @swanson1978             |
+| Social complexity                                     | @winkelman1986           |
+| Social flexibility, cognatic kinship                              | @shaara1992              |
+| Decreased social participation                                | @douglas2004natural      |
+| Social rigidity, monotheism                                                                               | @wood2018complexity      |
 
@@
 The data are challenging in two ways.
-The variables are themselves correlated, either because they measure different facets of the same thing (like how complex a society is) or because they were adapted together by socially or geographically close societies (Galton's problem).
+The variables are themselves correlated, either because they measure different facets of the same thing (like how complex a society is; see the figure below) or because they were adapted together by socially or geographically close societies (Galton's problem).
 In addition, data are missing from the Atlas in a non-random manner (e.g.
 if a society is hard to access, it is harder to provide a population size estimate for it).
+
+```{r }
+cord = wd %>% 
+  filter(in_sccs) %>% 
+  select(-outcome,-EA112_Trance_states,-trance_present,-possession_present,-possession_trance_present,-soc_id,-in_sccs) %>% 
+  mutate_if(is.logical, as.double)
+
+cor(cord,use="pairwise.complete.obs")
+
+```
 

### 2022-10-23 Readme.Rmd
fixed var coding, pls fix cor plot

@@
 ---
-
-# To do: one-hot encoding ended up creating unidentifiable vars (where both values of a col are present) so do something with that.
 
@@
 
-cor(cord,use="pairwise.complete.obs")
+# cor(cord,use="pairwise.complete.obs")
 
@@
 The last two columns show the error rate.
-The model's accuracy on absence of possession trance is relatively low, though higher than chance (54 out of 80, p < 0.01 using a binomial test with a true p of .529).
-It is mostly accurate capturing the presence of possession trance.
 
@@
 
-# precision = true 1/false 1
+# precision = true 1/tru1 + false 1
 # recall = true 1/true1 + false0
-precision = 68 / 94
-recall = 68 / 71
+precision = 64 / (64+12)
+recall = 64 / (64+7)
 f = (2 * precision * recall) / (precision + recall)
+chisq.test(c(c(68,12),c(7,64)))
 ```
 
-The model has an F-measure of `r round(f,2)`, with a penalty on overgeneralising possession trance.
+The model has an F-measure of `r round(f,2)`, with a penalty on slightly overgeneralising possession trance.
 Though comparisons are hard to make, this is not particularly accurate for a binary categorisation problem.
@@
 
-# precision = true 1/false 1
+# precision = true 1/true1+false 1
 # recall = true 1/true1 + false0
-precision2 = 221 / 389
-recall2 = 221 / 281
+precision2 = 194 / (194+113)
+recall2 = 194 / (194+87)
 f2 = (2 * precision2 * recall2) / (precision2 + recall2)
-
+chisq.test(c(c(264,113),c(87,194)))
 ```
@@
 They come in order of how important they are, according to the model; prevalence of slavery, then family organisation, and so on.
-The predictor levels are ordered according to coding and generally go from less prevalent (top) to more prevalent (bottom):
+The predictor levels are ordered according to coding and generally go from a pattern being less prevalent (top) to more prevalent (bottom):
 
-For the most important variables, we see a clear correllation between the presence of possession trance and the variable scaling up.
-Societies in which slavery is more prevalent have more possession trance than those where it is less prevalent.
-Same is true for societies with more complex family structures, more social complexity, more complex settlement patterns, and larger settlements.
-The societal variables that do not, strictly speaking, measure social organisation are less obvious.
+Ee see a clear correlation between the presence of possession trance and the variable scaling up, e.g.
+societies in which slavery is more prevalent have more possession trance than those where it is less prevalent.
+The across-level distributions are also interesting, e.g.
+the main distinction for patterns of slavery is for societies without slavery (first factor level) and all others with some form of slavery (subsequent factor levels).
 
@@
 @bourguignon1968 explicitly tests the variables of slavery, jurisdictional hierarchy, settlement patterns, and kin groups.
-@shaara1992 point to a high value placed on female virginity at marriage, captured here by norms of premarital sexual behaviour for girls.
+@shaara1992 point to a high value placed on female virginity at marriage, captured here by norms of premarital sexual behaviour for girls, @wood2018complexity point to the role of moralising high gods.
 These are all important variables in our model.
-@greenbaum1973 and @shaara1992 point to sexual division of labour, while @wood2018complexity to the role of moralising high gods, neither of which is relevant in our model.
+@greenbaum1973 and @shaara1992 point to sexual division of labour but this is not relevant in our model.
+
+Note that @shaara1992 use cognatic kinship as a proxy of social complexity (which in itself is robustly supported by our model) while the measure referenced by @wood2018complexity, the presence of moralising high gods, has in itself been argued to correlate with social complexity (see @botero2014ecology and @watts2015broad).
 
@@
 |more complexity$\rightarrow{}$trance w/ possession|@winkelman1986|yes, see above|
-|cognatic kinship$\rightarrow{}$more possession and trance|@shaara1992|not directly, but kinship structure is relevant|
+|cognatic kinship$\rightarrow{}$more possession and trance|@shaara1992|not directly|
 |decreased social participation$\rightarrow{}$possession|@douglas2004natural|yes, see above|
-|monotheism|@wood2018complexity|not relevant in our model|
+|monotheism|@wood2018complexity|yes|
 

### 2022-10-23 helper.R

@@ -75,7 +75,7 @@ dat = dat %>%
 # --- filter --- #
 
 # these are the variables of interest for trance and possession
-keep_vars = paste('EA', c("008","012","015","023","027","030","031","032","033","034","042","043","053","054","066","068","070","072","073","074","078","112","113"), sep='')
+keep_vars = paste('EA', c("008","012","015","023","030","031","032","033","034","042","043","053","054","066","068","070","072","073","074","078","112","113"), sep='')
 
 # we recode some of these as ordinal variables, others as factors, we leave the rest as is, thereby assuming it is an ordinal variable
 trance_dict = codes %>% 
@@ -94,12 +94,9 @@ trance_dict = codes %>%
         var_id == 'EA023' & code %in% c(1,2,3,4,5,6,9,13) ~ '3',
         var_id == 'EA023' & code %in% c(10) ~ '4',
         var_id == 'EA043' & code %in% c(1) ~ "patrilineal",
-        var_id == 'EA043' & code %in% c(2:7) ~ "other",
-        var_id == 'EA027' & code %in% c(4) ~ '1',
-        var_id == 'EA027' & code %in% c(3) ~ '2',
-        var_id == 'EA027' & code %in% c(5,8) ~ '3',
-        var_id == 'EA027' & code %in% c(1,6) ~ '4',
-        var_id == 'EA027' & code %in% c(7,2) ~ '5',
+        var_id == 'EA043' & code %in% c(3) ~ "matrilineal",
+        var_id == 'EA043' & code %in% c(2,5,6) ~ "cognatic",
+        var_id == 'EA043' & code %in% c(4,7) ~ "other",
         var_id == 'EA042' & code %in% c(7) ~ "int_agr",
         var_id == 'EA042' & code %in% c(5,6,9) ~ "ext_agr",
         var_id == 'EA042' & code %in% c(4) ~ "pastoralism",
@@ -195,49 +192,14 @@ datwo = datw %>%
   map_df(., ~ str_replace(., '^','l_'))
 
 datwc = datw %>% 
-  select(matches('EA(012|015|042|043|053|054|072|073|074|113)'))
+  select(matches('EA(012|015|042|043|053|054|072|073|074|113)')) %>% 
+  mutate_all(fct_infreq) # rerank these to show some consistent pattern: most populous level is highest
 
-# this is a HACK
-datwc = as_tibble(predict(dummyVars(" ~ .", data = datwc), newdata = datwc))
[…]

### 2022-10-23 model_gbm.R

@@ -82,7 +82,7 @@ gbm_grid1 = h2o.grid("gbm",
                      hyper_params = gbm_params1
 )
 
-h2o.saveGrid(grid_directory = 'grid', grid_id = 'gbm_grid1')
+# h2o.saveGrid(grid_directory = 'grid', grid_id = 'gbm_grid1')
 
 # pick the best models from grid
 gbm_grid_best1 = h2o.getGrid(grid_id = "gbm_grid1",

### 2022-10-27 Readme.Rmd
rewamp, pcm

@@
 library(patchwork)
+library(corrplot)
 
@@
 wd = read_tsv('data/dat_wide.tsv')
+bf = read_tsv('models/bayes_factors.tsv')
+cm1a = read_tsv('models/conf_matrix11model.tsv')
+cm1b = read_tsv('models/conf_matrix11test.tsv')
+varimp1 = read_tsv('models/varimp11.tsv')
 sccs = filter(ld, in_sccs)
@@
 
-We want to look at two questions.
+We want to look at three questions.
 What sort of social factors have been proposed to co-occur with trance and possession phenomena?
 Using the Ethnographic Atlas and the Standard Cross-Cultural Sample, can we get more confident that these factors do co-occur with such phenomena across cultures?
+Using phylogenetic comparative methods, can we make claims on the causative structure between the co-occurent factors?
 
@@
 
-Several works made the same point or emphasised various aspects of the same relationship. The above table is somewhat reductive.
+Several works made the same point or emphasised various aspects of the same relationship. The above table is somewhat reductive. The argument is spelt out in the main text.
 
@@
 if a society is hard to access, it is harder to provide a population size estimate for it).
+The correlations of the pairwise complete observations of the relevant predictors in the Ethnographic Atlas can be seen below.
 
 ```{r }
-cord = wd %>% 
-  filter(in_sccs) %>% 
-  select(-outcome,-EA112_Trance_states,-trance_present,-possession_present,-possession_trance_present,-soc_id,-in_sccs) %>% 
-  mutate_if(is.logical, as.double)
+cord = ld %>% 
+  filter(!var_title %in% c('Trance states','Societal rigidity')) %>% 
+  select(soc_id,var_title,code) %>%
+  pivot_wider(id_cols = soc_id, names_from = var_title, values_from = code) %>% 
+  select(-soc_id)
 
-# cor(cord,use="pairwise.complete.obs")
+cor(cord,use="pairwise.complete.obs") %>% 
+  corrplot()
 
@@
 
-The remaining co-variation and the non-random patterns of missing data recommends an ensemble learning method.
-This starts with societies with information on trance and possession phenomena and builds a lot of tiny models, using random samples of the data and random subsets of the covariates.
-Then, we compare these models to assess the overall importance of our covariates.
-Building little models in parallel is called bagging.
+The remaining co-variation and the non-random patterns of missing data recommend an ensemble learning method.
+This starts with societies that have information on trance and possession phenomena.
+It then builds a lot of tiny models, using random samples of the data and random subsets of the covariates.
+Then, we compare the tiny models with each other to assess the overall importance of our covariates.
+Building tiny models in parallel is called bagging.
 This is how a random forest works, growing many classification or regression trees and aggregating over the results.
@@
 
-Following @bourguignon1968, we reduce the complexity of the outcome variable to two levels: whether a society has possession trance or not.
+Following @bourguignon1968, we reduce the complexity of the outcome variable to two levels: whether a society has *possession trance* or not.
 The predictors are the possible covariates identified above.
 Some predictors, like `EA032` (Jurisdictional hierarchy of local community) or `EA113` can be construed to follow a scale.
-These are input as ordered factors.
-Others, like `EA053` (Sex differences in animal husbandry) or `EA043` (Major type of descent) can be carved up into categories: These are one-hot encoded.
+These are input as *ordered factors*.
+Others, like `EA053` (Sex differences in animal husbandry) or `EA043` (Major type of descent) can be carved up into categories: These are *one-hot encoded*.
 For details, see the Appendix.
 
-We use hyperparameter tuning to find the best possible model.
+We use *hyperparameter tuning* to find the best possible model.
 The model is fit on a training set of `length(unique(sccs$soc_id))` societies with information on trance and possession in the standard cross-cultural sample.
+The hyperparameter grid is printed in the Appendix.
 
@@
 
-We compare this with how our best model categorises each society.
+We compare training information on each society with how our best model categorises each society.
 This is our model's confusion matrix.
@@
 The last two columns show the error rate.
+For example, there are 71 societies in the Standard Cross-Cultural Sample that have possession trance, out of these, the model thinks 64 have possession trance and 7 do not (second row).
+This is an error rate of 7/71=.1.
 
 ```{r }
-cm1a = read_tsv('models/conf_matrix11model.tsv')
+
 cm1a %>% 
@@
 f = (2 * precision * recall) / (precision + recall)
-chisq.test(c(c(68,12),c(7,64)))
+# chisq.test(c(c(68,12),c(7,64)))
 ```
@@
 The model has an F-measure of `r round(f,2)`, with a penalty on slightly overgeneralising possession trance.
-Though comparisons are hard to make, this is not particularly accurate for a binary categorisation problem.
-We did expect this, though, as it is unlikely that any complex cultural phenomenon could be reduced to a small number of cross-cultural predictors.
+Though such comparisons are hard to make, this is not particularly accurate for a binary categorisation problem.
+This is not surprising -- it is unlikely that any complex cultural phenomenon could be reduced to a small number of cross-cultural predictors.
 
@@
 
-We can ask the model to make predictions for the entire Ethnographic Atlas (where trance and possession data are available) -- including the SCCS subset.
-Generally, you want to separate the training and test sets in Machine Learning and not do this, but the 'everything except the SCCS' subset of the Atlas would be an odd test set.
+We can ask the model to make predictions for the entire Ethnographic Atlas (where trance and possession data are available) -- including the Standard Cross-Cultural Sample subset.
+Generally, you want to separate the training and test sets in Machine Learning and not do this, but the 'everything except the Standard Cross-Cultural Sample' subset of the Atlas would be an odd test set.
 
@@
 ```{r }
-cm1b = read_tsv('models/conf_matrix11test.tsv')
 
@@
 Accuracy here is much worse than for the training data only.
-That is at least partly because the model is ignorant of the phylogenetic signal which might result in patterns that are unexpected from a purely correlational point of view.
+That is partly due to model bias and at least partly because the model is ignorant of the phylogenetic signal which might result in patterns that are unexpected from a purely correlational point of view.
 However, the model is still more accurate than it would be by chance.
 
-We take a look at the variables that were most important in the model in predicting the presence of possession trance in a given society, with a cutoff of 10/47.
+We now take a look at the variables that were most important in the model itself in predicting the presence of possession trance in a given society, with a cutoff of 10/47.
 
@@
 
-varimp1 = read_tsv('models/varimp11.tsv')
 vars2 = varimp1 %>% 
@@
 
-The figure below shows the proportion of societies with possession trance (dark blue) and without (light blue) across the ordered levels of the predictors that the model thinks are important.
+The figure below shows the proportion of societies with possession trance (dark blue) and without (light blue) across the ordered levels of the predictors.
+These are the predictors that the model thinks are important.
 They come in order of how important they are, according to the model; prevalence of slavery, then family organisation, and so on.
-The predictor levels are ordered according to coding and generally go from a pattern being less prevalent (top) to more prevalent (bottom):
+The predictor levels are ordered according to coding and generally go from a pattern being less complex or prevalent (top) to more complex or prevalent (bottom).
+So, for domestic organisation (top right) we go from small nuclear to large extended families.
+For slavery (top left) we go from absence of slavery to hereditary slavery.
 
-Ee see a clear correlation between the presence of possession trance and the variable scaling up, e.g.
-societies in which slavery is more prevalent have more possession trance than those where it is less prevalent.
+We see a clear correlation between the presence of possession trance and the variable scaling up.
+E.g.
+societies in which slavery is more prevalent are more likely to have possession trance than those where it is less prevalent.
+Societies with more complex extended families are more likely to have possession trance.
 The across-level distributions are also interesting, e.g.
@@
 
-## 7. Summary
+## 7. Did these traits develop independently?
+
+under construction.
+
+slavery. domestic organisation.
+binary.
+austronesian (eeh http://lir.byuh.edu/index.php/pacific/article/view/141)
+afro-asiatic (sub-saharan, greenbaum)
+atlantic-congo (hamito-semitic! lewis)
+
+bayestraits
+
+The independent model assumes the two traits evolve independently, e.g. the transition
+from 0 → 1 in the first trait is independent of the state of the second trait. 
+dependent: the traits are correlated and the rate of change in one
+trait is dependent on the state of the other
+100 stones and 1000
+
+mcmc exp 10 priors
+iterations per stone to estimate the marginal likelihood
+
+bayes factor
+```{r }
+kable(bf, 'simple', digits = 2)
+```
+
+lot of forking paths. slavery in atlantic-congo noteworthy. maybe. overall not really.
+
+## 8. Summary
 
@@
 Some of the important co-variates are the ones mentioned by earlier work and the direction of the relationships corroborates the literature.
+
+something something phylogenetic comparative methods
 
@@
 
+### Hyperparameter values for the Gradient Boosting Model
+
+```{r, echo=T}
+gbm_params1 = list(
+  learn_rate = c(0.01,0.02,0.03,0.1),
+  max_depth = c(1,3,5,7),
+  sample_rate = c(0.4,0.7,0.9,0.95),
+  col_sample_rate = c(0.1,0.2,0.3),
+  min_split_improvement = c(1e-3,1e-5)
+)
+```
+
 ## References

### 2022-10-27 helper.R

@@ -11,6 +11,7 @@ library(caret)
 library(rjson)
 library(RCurl)
 library(vctrs)
+library(ape)
 
 # --- pull --- #
 

### 2022-10-27 model_bayestraits.R

@@ -1,0 +1,153 @@
+setwd('~/Github/Racz2023trance/')
+library(tidyverse)
+library(glue)
+
+library(ape)
+library(ggtree)
+library(bayestraitr) # sam passmore's bayestraits helper library
+library(magrittr)
+
+# -- in -- #
+
+# data from helper.r
+dw = read_tsv('data/dat_wide.tsv')
+dl = read_tsv('data/dat_long.tsv')
+
+# trees from d-place
+tr_aut = read.nexus("https://raw.githubusercontent.com/D-PLACE/dplace-data/master/phylogenies/glottolog_aust1307/summary.trees")
+tr_afra = read.nexus("https://raw.githubusercontent.com/D-PLACE/dplace-data/master/phylogenies/glottolog_afro1255/summary.trees")
+tr_atla = read.nexus("https://raw.githubusercontent.com/D-PLACE/dplace-data/master/phylogenies/glottolog_atla1278/summary.trees")
+
+# -- formatting -- #
+
+dw %<>% 
+  mutate(
+    slavery_present = EA070_Slavery_type == "l_1",
+    domestic_organisation_nuclear = EA008_Domestic_organization %in% c('l_1','l_2')
+  )
+
+dw2 = dl %>% 
+  distinct(soc_id,society,glottocode,family) %>% 
+  left_join(dw) %>% 
+  select(family,glottocode,slavery_present,domestic_organisation_nuclear,possession_trance_present) %>% 
+  mutate_if(is.logical,as.double) %>% 
+  mutate_if(is.double,as.character)
+
+d_aut = dw2 %>% 
+  filter(family == 'Austronesian') %>% 
+  as.data.frame()
+d_afr = dw2 %>% 
+  filter(family == 'Afro-Asiatic',glottocode != 'kano1249') %>% 
+  as.data.frame()
+d_atl = dw2 %>% 
+  filter(family == 'Atlantic-Congo',glottocode != 'boro1274') %>% 
+  as.data.frame()
+
+rownames(d_aut) = d_aut$glottocode
+rownames(d_afr) = d_afr$glottocode
+rownames(d_atl) = d_atl$glottocode
+
+# -- bayestraits: out -- #
+
+bt_write(tree = tr_aut, data = d_aut, variables = c('slavery_present','possession_trance_present'), filename = 'Bayestraits/aut_sl')
[…]

### 2022-10-29 Readme.Rmd
finished readme

@@
 
-We want to look at three questions.
+This is the online supplement for the paper "Possession trance co-varies with measures of social rigidity in the Ethnographic Atlas 2022".
+The current version of the paper is [here](https://docs.google.com/document/d/1qbcsCeQ-Tw2W-yhQvvYd1Nk5ImQznqWBmiHX09lB_24/edit?usp=sharing).
+
+We have three questions.
 What sort of social factors have been proposed to co-occur with trance and possession phenomena?
-Using the Ethnographic Atlas and the Standard Cross-Cultural Sample, can we get more confident that these factors do co-occur with such phenomena across cultures?
-Using phylogenetic comparative methods, can we make claims on the causative structure between the co-occurent factors?
+Can we give support to these using the Ethnographic Atlas and the Standard Cross-Cultural Sample?
+Using phylogenetic comparative methods, can we make claims on the causative structure between the co-occurrent factors?
+
+Our informal priors are that trance and possession phenomena result from complex interchanges between indidvidual psychopathology and social organisation, and that, even if we can recognise broader patterns, the effects of social rigidity and marginalisation will be idiosyncratic on the ground.
+This means that both the outcome (trance and possession) and the predictors (e.g.
+measures of social rigidity) will be much more loosely defined than in similar analyses.
+So we expect any effects we find to be relatively modest.
 
@@
 
-Many people have argued that trance induced by possession is not a random, idiosyncratic pathology but a recurrent psycho-social trait that is at least partly motivated by other social factors.
+The paper says that trance induced by possession is not a random, idiosyncratic pathology but a recurrent psycho-social trait that is at least partly motivated by other social factors.
+This is recurrent claim in the literature.
+Here, we revisit the main points of this argument.
 
-High overall social complexity and social rigidity -- the resulting power dynamics -- might translate to possession trance practices.
+High overall social complexity and social rigidity, as well as the resulting power dynamics, might translate to possession trance practices.
 Such practices might serve to express or relay perceived social inflexibility and inequality.
 This is echoed in the medical anthropology literature (@during2011critical,@dein2012religion) and ties in with broader questions of inequality and mental health (@nguyen2003anthropology).
-Various causative models have been proposed to account for the emergence of possession trance and, specifically, shamanism in a social setting (@winkelman2015shamanism, @singh2018cultural, @wood2018trance).
+Various causative models have been proposed to account for the emergence of possession trance, shamanism, and witchcraft in a social setting (@winkelman2015shamanism, @singh2018cultural, @wood2018trance, @singh2021magic).
 
@@
 
-Several works made the same point or emphasised various aspects of the same relationship. The above table is somewhat reductive. The argument is spelt out in the main text.
+Several works made the same point or emphasised various aspects of the same relationship. The above table is somewhat reductive.
 
@@
 
-The Atlas records the presence of trance and possession phenomena in a given society, using the following variable (EA112):
+What is our outcome variable? The Atlas records the presence of trance and possession phenomena in a given society, using the following variable (EA112):
 
@@
 
-This is basically a truth table of whether possession beliefs or trance states are present in a given society, with some causal links between the two (if possession happens it leads to trance, if trance happens it's because of possession, etc).
-We can think of different groupings (trance phenomena encompass levels 1,3,4,5,6,7, possession phenomena encompass 2,3,4,5,6,7, specifically possession trance covers 3,4,6,7 but not 1,2,5,8).
+This is basically a truth table of whether possession beliefs or trance states are present in a given society, with some causal links between the two (if possession happens it leads to trance, if trance happens it is because of possession, etc.).
+We can think of different groupings: trance phenomena encompass levels 1,3,4,5,6,7, possession phenomena encompass 2,3,4,5,6,7, specifically possession trance covers 3,4,6,7 but not 1,2,5,8.
+Following @bourguignon1968, we reduce the complexity of the outcome variable to two levels: whether a society has *possession trance* or not.
 
+```{r }
+trance %>% 
+  group_by(possession_trance) %>% 
+  summarise(count = sum(`n in SCCS`)) %>% 
+  kable('simple')
+```
+
+What are our predictor variables?
 The Atlas lists a set of variables that tie in with the observed covariates in the literature.
@@
 
-## 3. How do we look for correlations between trance and possession and their proposed societal covariates?
+## 3. How do we look for correlations between possession trance and its proposed societal covariates?
 
@@
 
-Following @bourguignon1968, we reduce the complexity of the outcome variable to two levels: whether a society has *possession trance* or not.
 The predictors are the possible covariates identified above.
@@
 From the rest, a majority have possession trance in some form.
-Following @bourguignon1968 we can reduce this to a binary: 81 societies have possession trance, 70 have other forms of trance and possessior or show neither.
-
-```{r }
-trance %>% 
-  group_by(possession_trance) %>% 
-  summarise(count = sum(`n in SCCS`)) %>% 
-  kable('simple')
-```
+Following @bourguignon1968 we can reduce this to a binary: 81 societies have possession trance, 70 have other forms of trance and possession or show neither.
 
@@
 
-We now take a look at the variables that were most important in the model itself in predicting the presence of possession trance in a given society, with a cutoff of 10/47.
+We now take a look at the variables that were important in the model itself in predicting the presence of possession trance in a given society, with a cutoff of 10/47 most important variables.
 
@@
 
-We now look at the direction of these effects (does e.g.\ increased jurisdictional hierarchy go with more or less prevalent possession trance?).
+Before we cross-reference the results with the claims in the literature, we look at the direction of these effects (does e.g.\ increased jurisdictional hierarchy go with more or less prevalent possession trance?).
+Then we ask questions about co-occurrence versus co-evolution.
 
@@
 
-under construction.
+The literature makes causative claims: societal rigidity causes possession trance (and not the other way round).
+Cross-cultural correlations can be indicative of causative structure but cannot provide direct evidence for it.
 
-slavery. domestic organisation.
-binary.
-austronesian (eeh http://lir.byuh.edu/index.php/pacific/article/view/141)
-afro-asiatic (sub-saharan, greenbaum)
-atlantic-congo (hamito-semitic! lewis)
+In addition, we can look at the cultural evolution of the outcome and some of the predictor variables.
+We take two examples, slavery and domestic organisation.
+If the practice of slavery leads to possession trance, we expect them to (i) correlate in the ethnographic record (they do) and emerge together in cultural phylogenies.
 
-bayestraits
+To test for this, we select two predictors, slavery and domestic organisation.
+These are the two strongest predictors in our cross-cultural analysis.
+We recode them as binaries to reflect their relationship to prevalence of possession trance: For prevalence of slavery, 1 is any form of slavery present in a cultural group, 0 is no slavery attested.
+For domestic organisation, 1 is nuclear family organisation, 0 is larger extended family organisation.
+Our outcome is 1 where possession trance is present and 0 where it is absent.
 
-The independent model assumes the two traits evolve independently, e.g. the transition
-from 0 → 1 in the first trait is independent of the state of the second trait. 
-dependent: the traits are correlated and the rate of change in one
-trait is dependent on the state of the other
-100 stones and 1000
+We select three phylogenies, Austronesian, Afro-Asiatic, and Atlantic-Congo.
+These are robust phylogenies with a large number of societies in each.
+They include societies that have been discussed by the trance and possesion literature (@blust2000rat, @greenbaum1973, @boddy1989wombs, @lewis1991introduction).
 
-mcmc exp 10 priors
-iterations per stone to estimate the marginal likelihood
+We use BayesTraits to assess the co-evolution of predictors and outcome (@pagel2007bayestraits).
+We fit two models, an independent model, which assumes that two traits evolve independently, and a dependent model, which assumes that traits are correlated and the rate of change in one trait.
+is dependent on the state of the other trait.
+
+Six dependent and six independent models are fit, for the relationship of (i) possession trance and the presence of slavery, (ii) possession trance and nuclear/extended family organisation across the three phylogenies.
+
+Phylogenies come from D-Place.
+Societies are matched to phylogenetic trees by language ID-s.
+One language is used by more than one society in Afro-Asiatic and in Atlantic-Congo respectively, these are removed.
+Societies that have no data on both traits are pruned from the phylogenetic trees.
+
+Models are fit using Markov Chain Monte Carlo using exp 10 priors, 100 stepping stones, and 1000 iterations per stone. Each run provides a marginal likelihood for the model fit.
+
+We calculate the Bayes Factors for the more complex dependent versus the less complex independent models.
+A higher value for a Bayes Factor indicates that the more complex model provides a better fit and so its complexity is justified.
+The Bayes Factors for the six model comparisons can be seen below.
 
@@
 
-lot of forking paths. slavery in atlantic-congo noteworthy. maybe. overall not really.
+The results strongly suggest that possession trance and slavery as well as nuclear family organisation developed independently across the three phylogenies.
+One apparent exception is slavery and possession trance in Afro-Asiatic languages, which of course cover the famed description of possession trance practices in the Sudan and the Horn of Africa.
+But a Bayes Factor of 4 is not substantial.
+Overall, there is little evidence here for trait co-evolution.
+
+The phylogenetic comparative analysis concedes with a large number of limitations. It remains true that it does not lend additional support to the cross-cultural analysis.
 
@@
 
-These results go some way in confirming hypotheses on the cross-cultural correlates of possession trance in the literature.
+The overall results go some way in confirming hypotheses on the cross-cultural correlates of possession trance in the literature.
 @bourguignon1968 explicitly tests the variables of slavery, jurisdictional hierarchy, settlement patterns, and kin groups.
 @shaara1992 point to a high value placed on female virginity at marriage, captured here by norms of premarital sexual behaviour for girls, @wood2018complexity point to the role of moralising high gods.
-These are all important variables in our model.
+These are all important variables in our cross-cultural model.
 @greenbaum1973 and @shaara1992 point to sexual division of labour but this is not relevant in our model.
@@
 
-We can predict some types of trance and possession phenomena in the SCCS subset of the EA.
+We can predict some types of trance and possession phenomena in the Standard Cross-Cultural Sample subset of the Ethnographic Atlas.
 Not all of them.
 Some of the important co-variates are the ones mentioned by earlier work and the direction of the relationships corroborates the literature.
-
-something something phylogenetic comparative methods
+Phylogenetic comparative methods do not provide additional support to attested cross-cultural co-variation.
 
 Human society is complicated and cross-cultural data have very low resolution.
-In this sense, we'd be very suspect of a model that would be too accurate of predicting possession trance from societal covariates.
-It cannot replace fieldwork and more complex explanatory models either.
-What it can do is provide an extra leg for such explanatory work, which is broadly the recognised ambition of cross-cultural comparative work in evolutionary anthropology.
+In this sense, we'd be suspect of a model that would be too accurate of predicting possession trance from societal covariates.
+It cannot replace fieldwork and more complex explanatory models.
+It can provide an extra leg for such explanatory work, which is the recognised ambition of cross-cultural comparative work in evolutionary anthropology.
 
 Our approach is statistically strict, so the fact that it still shows some accuracy is definitely promising for the validity of the connections drawn by previous work on trance and possession phenomena in particular and large cross-cultural comparisons in general.
+
+## 9. Limitations
+
+The cross-cultural model is fit on the Standard Cross-Cultural Sample which does not minimise the phylogenetic signal completely (see [here](https://excd.org/2018/03/01/stats-corner-is-the-standard-cross-cultural-sample-really-standard/)).
+The data in the Ethnographic Atlas cannot be independently verified.
+The outcome variable and the predictors are selected and defined with the existing literature in mind.
+It remains true that these decisions use up a large amount of researcher degrees of freedom.
+This is especially true for the phylogenetic comparative analysis, which further reduces resolution on the co-varying traits as a starting point.
+
+Such limitations will be in place for all cross-cultural analyses.
+They are exacerbated here by the elusiveness of the variables themselves, which likely results in more starting variability baked into the Ethnographic Atlas coding in the first place.
+The accompanying statistical rigour might render this work relevant to the broader study of trance and possession, despite these limitations.
 

### 2022-10-29 Readme.Rmd
now is good

@@
 
-This is the online supplement for the paper "Possession trance co-varies with measures of social rigidity in the Ethnographic Atlas 2022".
-The current version of the paper is [here](https://docs.google.com/document/d/1qbcsCeQ-Tw2W-yhQvvYd1Nk5ImQznqWBmiHX09lB_24/edit?usp=sharing).
+# Online supplement: Possession trance co-varies with measures of social rigidity in the Ethnographic Atlas
+
+This is the online supplement for the paper “Possession trance co-varies
+with measures of social rigidity in the Ethnographic Atlas 2022”. The
+current version of the paper is
+[here](https://docs.google.com/document/d/1qbcsCeQ-Tw2W-yhQvvYd1Nk5ImQznqWBmiHX09lB_24/edit?usp=sharing).
+The `helper` file creates the data files.
+The `model` files run the Gradient Boosting Model and the Bayestraits model, respectively.
 
@@
 
-Our informal priors are that trance and possession phenomena result from complex interchanges between indidvidual psychopathology and social organisation, and that, even if we can recognise broader patterns, the effects of social rigidity and marginalisation will be idiosyncratic on the ground.
+Ee expect any effects we find to be relatively modest.
+This is because trance and possession phenomena result from complex interchanges between individual psychopathology and social organisation, and that, even if we can recognise broader patterns, the effects of social rigidity and marginalisation will be idiosyncratic on the ground.
 This means that both the outcome (trance and possession) and the predictors (e.g.
-measures of social rigidity) will be much more loosely defined than in similar analyses.
-So we expect any effects we find to be relatively modest.
+measures of social rigidity) will be much more inherently variable than in similar analyses.
 
@@
 This is recurrent claim in the literature.
-Here, we revisit the main points of this argument.
+Here, we revisit the main points of this claim.
 
@@
 Several works made the same point or emphasised various aspects of the same relationship. The above table is somewhat reductive.
+
+Broadly speaking, we have measures of:
+
+1. Social complexity (including monotheism, which we assume to be more prevalent in more complex societies, see @botero2014ecology)
+2. Social rigidity (including cognatic kinship, which @shaara1992 uses as an example of a more flexible social blueprint)
+3. Social participation, with an emphasis on gender roles in social activites, including work.
 
@@
 
-```{r }
-ld %>% 
+Variables 8, 30:33, 66, 68, 70, 72, 73 were lifted from previous work, most notably @bourguignon1968 and @greenbaum1973.
+In addition, Variables 15, 23, 27, 74, 113 were taken as approximations for social rigidity, while variables 53, 54, 78 for participation, based on the literature above.
+
+```{r eval = F}
+pred_table = ld %>% 
   filter(var_title != 'Trance states') %>% 
@@
   arrange(var_id) %>% 
-  rename(ID = var_id, Name = var_title, Definition = var_definition) %>% 
+  rename(ID = var_id, Name = var_title, Definition = var_definition)
+
+write_tsv(pred_table, 'data/pred_table_temp.tsv')
+```
+
+```{r }
+pred_table = read_tsv('data/pred_table_added.tsv')
+pred_table %>% 
   kable('simple')
@@
 We use *hyperparameter tuning* to find the best possible model.
-The model is fit on a training set of `length(unique(sccs$soc_id))` societies with information on trance and possession in the standard cross-cultural sample.
+The model is fit on a training set of `r length(unique(sccs$soc_id))` societies with information on trance and possession in the standard cross-cultural sample.
 The hyperparameter grid is printed in the Appendix.
@@
 The overall results go some way in confirming hypotheses on the cross-cultural correlates of possession trance in the literature.
-@bourguignon1968 explicitly tests the variables of slavery, jurisdictional hierarchy, settlement patterns, and kin groups.
+@bourguignon1968 explicitly tests the variables of domestic organisation slavery, jurisdictional hierarchy, settlement patterns, and kin groups.
 @shaara1992 point to a high value placed on female virginity at marriage, captured here by norms of premarital sexual behaviour for girls, @wood2018complexity point to the role of moralising high gods.

### 2022-10-30 Readme.Rmd
new figure

@@
 library(corrplot)
+library(ggthemes)
+library(ggfortify)
+library(ggmap)
+library(googlesheets4)
+
+mapWorld = map_data('world', wrap=c(-25,335), ylim=c(-55,75)) # pacific centered
 
@@
 measures of social rigidity) will be much more inherently variable than in similar analyses.
+
+Here is a map of trance and possession phenomena.
+
+```{r fig.width=18, fig.height=5}
+lld = ld %>% 
+  filter(var_id == 'EA112') %>% 
+  mutate(
+    lon2 = ifelse(lon < -20, lon + 360, lon),
+    var_short = case_when(
+      code == 1 ~ 'T',
+      code == 2 ~ 'P',
+      code == 3 ~ 'T,P,T iff P',
+      code == 4 ~ 'T,P,T if P',
+      code == 5 ~ 'T,P,T and P',
+      code == 6 ~ 'P,T if P',
+      code == 7 ~ 'T,P if T',
+      code == 8 ~ 'none'
+    ),
+    coding = fct_reorder(var_name,code)
+  )
+
+p1 = ggplot() + 
+  geom_polygon(data = mapWorld, aes(x=long, y = lat, group = group)) + 
+  geom_label(data = lld, aes(x = lon2, y = lat, colour = coding, label = code)) +
+  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks=element_blank(), axis.title.y=element_blank(), axis.text.y=element_blank()) +
+  scale_colour_viridis_d() +
+  theme_void()
+
+p1 + plot_layout(guides = 'collect')
+```
 
@@
 
-ld %>% 
+table1 = ld %>% 
   filter(var_title == 'Trance states', !is.na(code)) %>% 
@@
   rename(Coding = code, Description = var_description) %>% 
-  select(Coding,Description,possession_trance,`n in EA`,`n in SCCS`) %>% 
+  select(Coding,Description,possession_trance,`n in EA`,`n in SCCS`)
+
+#write_sheet(table1, ss = 'https://docs.google.com/spreadsheets/d/1X7ZCnw6LVhjLIy4SiSeK3iZ6o1S144iHMv4DqOaCDEU/edit#gid=0', sheet = 'Table 1')
+
+table1 %>% 
   kable('simple')
@@
 pred_table = read_tsv('data/pred_table_added.tsv')
+
+#write_sheet(pred_table, ss = 'https://docs.google.com/spreadsheets/d/1X7ZCnw6LVhjLIy4SiSeK3iZ6o1S144iHMv4DqOaCDEU/edit#gid=0', sheet = 'Table 3')
+
 pred_table %>% 
@@
 
-cm1a %>% 
+table4 = cm1a %>% 
   mutate(original = c(names(cm1a)[!names(cm1a) %in% c('Error','Rate')],'Total')) %>% 
-  relocate(original) %>% 
+  relocate(original)
+
+#write_sheet(table4, ss = 'https://docs.google.com/spreadsheets/d/1X7ZCnw6LVhjLIy4SiSeK3iZ6o1S144iHMv4DqOaCDEU/edit#gid=0', sheet = 'Table 4')
+
+table4 %>% 
   kable('simple', digits = 2)
@@
   kable('simple')
+
+#write_sheet(cm1b, ss = 'https://docs.google.com/spreadsheets/d/1X7ZCnw6LVhjLIy4SiSeK3iZ6o1S144iHMv4DqOaCDEU/edit#gid=0', sheet = 'Table 5')
 
@@
   select(ID, Name, relative_importance)
+
+#write_sheet(vars2, ss = 'https://docs.google.com/spreadsheets/d/1X7ZCnw6LVhjLIy4SiSeK3iZ6o1S144iHMv4DqOaCDEU/edit#gid=0', sheet = 'Table 6')
 
@@
 ```{r }
+#write_sheet(bf, ss = 'https://docs.google.com/spreadsheets/d/1X7ZCnw6LVhjLIy4SiSeK3iZ6o1S144iHMv4DqOaCDEU/edit#gid=0', sheet = 'Table 7')
 kable(bf, 'simple', digits = 2)

### 2023-02-01 Readme.Rmd
tidyup

@@
 ---
-title: Identifying covariate traits of trance and possession phenomena in the Ethnographic
-  Atlas using boosting and nested regression
+title: Supplementary Information, Possession trance covaries with measures of social rigidity in the Ethnographic Atlas 
 author: "Péter Rácz"
@@
 
-This is the online supplement for the paper “Possession trance co-varies
-with measures of social rigidity in the Ethnographic Atlas 2022”. The
+This is the online supplement for the paper “Possession trance covaries
+with measures of social rigidity in the Ethnographic Atlas”. The
 current version of the paper is
@@
 We have three questions.
-What sort of social factors have been proposed to co-occur with trance and possession phenomena?
+What sort of social factors have been proposed to cooccur with trance and possession phenomena?
 Can we give support to these using the Ethnographic Atlas and the Standard Cross-Cultural Sample?
-Using phylogenetic comparative methods, can we make claims on the causative structure between the co-occurrent factors?
+Using phylogenetic comparative methods, can we make claims on the causative structure between the cooccurrent factors?
 
-Ee expect any effects we find to be relatively modest.
+We expect any effects we find to be relatively modest.
 This is because trance and possession phenomena result from complex interchanges between individual psychopathology and social organisation, and that, even if we can recognise broader patterns, the effects of social rigidity and marginalisation will be idiosyncratic on the ground.
@@
 
-```{r fig.width=18, fig.height=5}
+```{r fig.width=8, fig.height=9}
 lld = ld %>% 
@@
     var_short = case_when(
-      code == 1 ~ 'T',
-      code == 2 ~ 'P',
-      code == 3 ~ 'T,P,T iff P',
-      code == 4 ~ 'T,P,T if P',
-      code == 5 ~ 'T,P,T and P',
-      code == 6 ~ 'P,T if P',
-      code == 7 ~ 'T,P if T',
+      code == 1 ~ 'trance',
+      code == 2 ~ 'possession',
+      code == 3 ~ 'trance, possession,\ntrance iff possession',
+      code == 4 ~ 'trance, possession,\ntrance if possession',
+      code == 5 ~ 'trance, possession,\ntrance and possession',
+      code == 6 ~ 'possession,\ntrance if possession',
+      code == 7 ~ 'trance,\npossession if trance',
       code == 8 ~ 'none'
     ),
-    coding = fct_reorder(var_name,code)
+    coding = fct_reorder(var_name,code),
+    var_short = fct_reorder(var_short,code)
   )
@@
   geom_polygon(data = mapWorld, aes(x=long, y = lat, group = group)) + 
-  geom_label(data = lld, aes(x = lon2, y = lat, colour = coding, label = code)) +
+  # geom_label(data = lld, aes(x = lon2, y = lat, label = ' ')) +
+  geom_point(data = lld, aes(x = lon2, y = lat), colour = '#FFAA00') +
   theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks=element_blank(), axis.title.y=element_blank(), axis.text.y=element_blank()) +
-  scale_colour_viridis_d() +
-  theme_void()
+  # scale_colour_viridis_d() +
+  # guides(colour = 'none',label = 'none') +
+  theme_void() +
+  facet_wrap( ~ var_short, nrow = 4)
 
-p1 + plot_layout(guides = 'collect')
+p1
 ```
@@
 
-Models are fit using Markov Chain Monte Carlo using exp 10 priors, 100 stepping stones, and 1000 iterations per stone. Each run provides a marginal likelihood for the model fit.
+Models are fit using Markov Chain Monte Carlo using exp(10) priors, 100 stepping stones, and 1000 iterations per stone. Each run provides a marginal likelihood for the model fit.
 
@@
 ```{r }
-ld %>% 
+coding_table = ld %>% 
   filter(var_id != 'EA112') %>% 
   distinct(var_id,var_title,var_description,recoded_value) %>% 
-  arrange(var_id,recoded_value) %>% 
+  arrange(var_id,recoded_value)
+
+coding_table %>% 
   kable('simple')
+
+# write_sheet(coding_table, ss = 'https://docs.google.com/spreadsheets/d/1X7ZCnw6LVhjLIy4SiSeK3iZ6o1S144iHMv4DqOaCDEU/edit#gid=0', sheet = 'Table 9')
+
 ```
@@
 )
+
 ```

### 2023-05-14 response_to_reviews.txt
new final draft

> todo
> read this
> read paper
> read rmd
> compile
> push to github; put to do dropbox

We would like to thank the Editor and the Reviewers. Reviewer comments made it clear that (i) we needed to make some of the epistemological uncertainty surrounding "social rigidity" and "possession trance" as unitary social phenomena more explicit, (ii) we needed to better justify the boosting model used in the analysis, and (iii) the phylogenetic comparative analysis was badly implemented and had to be rerun from scratch.

In response, we made considerable edits in the text and in the analyses.

Following suggestions by both reviewers, and, in particular, Reviewer 2, we now spend more time in both the introduction and the discussion on the following question: Whether various measures of social rigidity, complexity, etc. all approximate some abstract standard of inflexibility or gauge different social structures and, correspondingly, whether we can talk about a singular phenomenon of possession trance or a rather a range of related phenomena. Based on cross-cultural data, the answer to these questions is at the very least inconclusive (and more likely a resounding no) -- but this has to be made explicit in the paper itself.

Following suggestions by both reviewers, we now add more details to how the boosting model was implemented and why it was trained on the Standard Cross-Cultural Sample and tested on the Ethnographic Atlas. In order to provide further validation for the model, validation that the reviewers seemed to find welcome, we added a regression model analysis that explicitly controls for geographic and cultural co-variation. We also discuss the benefits and hazards of using such a model versus our boosting approach when working with cross-cultural data of this kind.

Following pointers by Reviewer 2, we repeated the phylogenetic comparative analysis using appropriate standards of model fitting and reporting.

Reviewer 1

Reviewer: I wonder more could be done to mitigate some of the limitations. For example, rather than using the SCCS because it tries to (but does not completely eliminate, as mentioned in the discussion) phylogenetic autocorrelation, why not build a mode that explicitly incorporates phylogenetic co-variance? See Claessens & Atkinson (2022) for examples of how to incorporates this into a regression model (I know that authors here use a decision tree approach, so the models are not directly comparable).

Response: It is clear that the modelling results require more framing and more discussion. We decided to add these in the new draft as well as include a regression analysis to explore other ways in which we could account for autocorrelation. Please find more details below.

Reviewer: The models used, I am not particularly familiar with. I think more contextualising of the model results would help readers understand the contexts further. For example: There is a section on model accuracy. This is an admirable inclusion, and I think perhaps something more people should include. However, there is no way for the reader to determine whether (e.g.) an F-measure of 0.66 is good or not. What level of accuracy should we expect? Is there some independent metric we can call on (like BayesFactors used in the subsequent analyses), or a comparison we can make to understand how well we should expect the model to perform?

Response: We fit a chi squared test of independence on the model's confusion matrix on the training and test data and report the results in the new draft. The chi squared test is more confident on the test data as there are more of these, but it is, broadly speaking, helpful in giving a noise estimate to accompany the signal estimate given by the F measure, precisely as the Reviewer suggested.

Reviewer: A second point in this model for clarification is that the parameters are only derived from the SCCS – is that correct? That is, the results in Table 5 are how well the SCCS model predicts the presence/absence of possession trance in the Ethnographic atlas – but the EA data has no bearing on the parameters of this model (other than the SCCS being a subset of the EA). A similar problem arises with the variable importance. Since the author(s) only perform analyses on the top two most important variables, should we infer that a Relative Importance value of less than 50 is bad? Or is the decision to only analyse the top two variables not driven by data?

Response: The Reviewer is correct, the entire EA dataset is only used to test the boosting model. We now make this clearer at the beginning of the modelling section. We didn't, as such, perform additional analysis on the ten most important variables in the original draft. It is true, however, that we only report the ten most important ones. The 10/47 cutoff in variable importance is arbitrary, although the ten most important variables already account for 91% of the variation explained by the model. We now expand on this detail in the Modelling section. We now also talk more about our choice of a model. The trade-off, as we see it, between a lot-of-tiny-models approach (like our many trees model) and a one-big-model approach (like a regression model) is that a tiny models approach is much better at using all data available, while a big model is much better at handling autocorrelation. We now make this more explicit in our Limitations section. In addition, in the Results section, we now include a validation analysis that uses Bayesian generalized linear regression, following pointers in Claessens & Atkinson (2022). This analysis largely supports the findings of our tiny-models approach and also provides an illustration of why it is difficult to use one large model that has many predictors on a heterogeneous dataset with systematically missing data.

Reviewer: My biggest concern in this paper is the phylogenetic data used in the co-evolution analyses. Three language families are listed: Austronesian, Afro-Asiatic, and Atlantic-Congo. The paper references that these come from D-PLACE, but it would be more appropriate to cite the original source of the phylogenies. My concern is that the authors have used Glottolog taxonomies in their phylogenetic analyses. This is not appropriate because taxonomies have not estimated branch-lengths for the traits to evolve down. You can see within the phylogenetic data files provided that all branch lengths are equal to 1. If the authors think they are justified in using taxonomic data instead of phylogenetic data, then they will need to justify that in the text, but I am unaware of any such argument. My current position is that this analysis is currently unpublishable.

Response: It is safe to say that our grasp on the co-evolution analysis was tenuous. The Reviewer's very specific feedback has ben extremely helpful in rewriting this section. Instead of Glottolog taxonomies, we now use two phylogenies, Austronesian and Atlantic-Congo, as there is no readily available phylogeny for Afro-Asiatic. We cite the original sources of these phylogenies beside D-Place.

Reviewer: In addition, more information is needed about the Bayestraits analyses. At minimum, we need: How many chains were run? Did the chains converge? How many iterations were there? What was the sampling rate? What were the priors? See Watts et al. for a good example of Bayestraits model description. Looking at the Log output for the models provided, we can see that the models are not performing particularly well, since the ancestral states are all 0.25 in many iterations.

Watts, J., Sheehan, O., Atkinson, Q. D., Bulbulia, J., & Gray, R. D. (2016). Ritual human sacrifice promoted and sustained the evolution of stratified societies. Nature, 532(7598), 228–231. https://doi.org/10.1038/nature17159

Response: In the reanalysis, we report on the number of chains run, number of iterations, sampling rate, priors, and convergence, beyond the Bayes Factors for the hypothesis pairs. We follow the Reviewer's pointers and the example of the Watts paper. The results of the reanalysis are, admittedly, inconclusive, but they could provide a valuable stepping stone for further work, and so are included in the new draft.

Reviewer: Some specific comments as I go through:

-       Figure 1: It would be helpful to label the maps with respect to the numbers in the caption. I find the titles difficult to follow.

Response: We replaced the captions with more informative ones and added numbering.

-       Figure 3: Put colour in the caption or in a legend, rather than a written caption.

Response: Fixed.

-       P17 L17 says that Figure 3 contains proportions, but that cannot be true. The slavery graph has an x-axis that goes to 200.

Response: This was wrong. Corrected to "counts".

-       More information of Bayestraits runs
o       How many chains, do the chain converge etc.
o       How many tree samples are used?

Response:  We have reran the analyses using phylogenetic trees instead of glottolog taxonomies and now include this information in the paper.

-       Bayestraits manual lists a BF of > 2 as positive and > 5 as strong, so a BF of 4 seems to be of note?

Response: This is a good point. In our new analysis, we have one model comparison where a BF of four supports the more complex model, and we do speculate on whether this is due to cultural factors.

-       The website linked to Racz (2018) does not exist: https://excd.org/2018/03/01/stats-corner-is-the-standard-cross-cultural-sample-really-standard/R%E1cz

Response: We replaced this with Eff 2004 which is a more exhaustive treatment of the same issue.

-       Although you mention the EA is low resolution,  this paper suggests does contain robust codings.

Bahrami-Rad, D., Becker, A., & Henrich, J. (2021). Tabulated nonsense? Testing the validity of the Ethnographic Atlas. Economics Letters, 204, 109880. https://doi.org/10.1016/j.econlet.2021.109880

Response: This is a brilliant and, to us, very reassuring reference. We added it in our Limitations section (i.e. "the EA might be a bit coarse but probably not too bad").

-       This isn’t a critique, but an interesting output of the independent model that isn’t explored is whether societies tend to gain or lose possession trance in any particular order.
o For example: once trance arises does it stay or can possession trance come and go?
o This information is in the Log file of the Bayestraits output. Although, given the comments above, the results are not currently interpretable.

Response: We agree with the Reviewer that this is an interesting path to follow up on. We also think that there is potential in a research paper on the cultural evolution of possession trance. However, we would rather not explore more of this angle in this paper. We feel that we lack the necessary expertise in comparative methods and sufficient experience with either Bantu or Austronesian to give a meaningful intepretation of any results beyond whether traits evolved independently. We are entirely sympathetic to the argument that the phylogenetic analysis in the paper should be either further extended (beyond entailing sufficient reporting detail) or dropped altogther, but we also think that, in its present form, it remains an intriguing addition to the argument without entirely taking over the argument altogether.

Reviewer: 2

Comments to the Author
Reviewer: For such an exhaustive piece of work I have little to add/suggest- generally this is very clearly described, but some sections could be expanded or expressed in less technical language.

Reviewer: You say that possession trance with social rigidity measures in the title, and then in the abstract you say the association is non-specific(i.e seems general to all social measures you looked at)- these are slightly inconsistent messages I would say. A response might be to say that actually these were not actually non-specific, but rather reflect that the predictors are all measures of social complexity. That is, how do you theorise the relationship between the measures and social complexity- are these distinct factors, or reflections of a single underlying factor. I think in places your report is consistent with each of these in different places.

Response: This is the million dollar question! Maybe it's possible to define a unified measure of social inflexibility and this will always make it more likely that possession trance emerges in a society. Maybe these are what they are: disparate measures of different things that might give rise to different types of trance and possession phenomena. We don't have a good answer here and are not sure if it's possible to give one within the confines of this cross-cultural paper. However, the question itself should be made explicit! We add a paragraph discussing it at the end of the Introduction and briefly speculate on it in the Discussion.

Reviewer: I would refer to the machine learning method in the abstract, given ML is a keyword.

Response: Added.

Review: Could it be better to moderate the first sentence from "comes from the presence" to "seems to come..."? This is consistent with your use of phenomena in the first line f the background, that PT is something which is experienced.

Response: We agree. Fixed.

Reviewer: Could say more about the Seligman 2005 reference on page 4 of the pdf?

Response: Seligman is a very nice example of how a pattern can be recontextualised from psychopathological condition to religious practice and so it is an ideal step forward from strictly biological or pathological framings of possession trance. We now discuss this in more detail in the text.

Reviewer: I think there might be a word missing here:

"These have been largely downplayed by previous work, despite that they can all lead to spurious correlations (Gelman & Hill, 2006)."

Response: Fixed.

Reviewer: I don't think I have come across "truth table" as a terminology before.

Response: Fixed.

Reviewer: Could you expand what is meant by one-hot coding in the MS?

Response: We added an example.

Reviewer: Could you clarify what the purpose of the validation of the model in the EA?

Response: Testing on the Atlas is a straightforward way of introducing the model to previously unseen data and to test its estimates on a larger dataset. Broadly speaking, it lends a lot more credence to the results we garnered from the Sample data. We now make this clear in the text.

Reviewer: Could you clarify what the scale is for relative importance? Is it e.g. a measure of magnitude for coefficients, or precision?

Response: Relative importance, broadly speaking, expresses the rate with which the inclusion of the variable decreases error in the model fitting process. This means that, in theory, a small but robust effect can be as important as a larger but noisier effect. We now make this clear in the text.

Reviewer: At Figure 3- please could you label the bars somewhere?

Response: Added label.

Reviewer: The co-evolution hypothesis is well tested and described, but does not seem well integrated into the Background and Introduction. Could you say something about why answering the second question is important?

Response: The literature makes claims on social rigidity causing possession trance (to put it fairly bluntly). The absence of clear-cut coevolution patterns means that strong, causative claims regarding the effect of social rigidity on the emergence of possession trance do not straightforwardly translate to a cultural evolution account of the phenomenon. This, together with the epistemological uncertainties pointed out by this Reviewer in their first comment, mean that a strong reading of possession trance as a unitary phenomenon caused by very specific social effects is probably untenable.

### 2023-05-14 Readme.Rmd
new final draft

@@
 library(googlesheets4)
+library(broom)
 
@@
 varimp1 = read_tsv('models/varimp11.tsv')
+glm1 = read_tsv('models/glm_estimates.tsv')
 sccs = filter(ld, in_sccs)
@@
 
-```{r fig.width=8, fig.height=9}
+```{r fig.width=8, fig.height=11}
 lld = ld %>% 
@@
     lon2 = ifelse(lon < -20, lon + 360, lon),
-    var_short = case_when(
-      code == 1 ~ 'trance',
-      code == 2 ~ 'possession',
-      code == 3 ~ 'trance, possession,\ntrance iff possession',
-      code == 4 ~ 'trance, possession,\ntrance if possession',
-      code == 5 ~ 'trance, possession,\ntrance and possession',
-      code == 6 ~ 'possession,\ntrance if possession',
-      code == 7 ~ 'trance,\npossession if trance',
-      code == 8 ~ 'none'
-    ),
     coding = fct_reorder(var_name,code),
-    var_short = fct_reorder(var_short,code)
+    var_description2 = glue('({code}) {var_description}')
   )
@@
   theme_void() +
-  facet_wrap( ~ var_short, nrow = 4)
+  facet_wrap( ~ var_description2, nrow = 4, labeller = label_wrap_gen(width = 60, multi_line = T))
 
@@
 f = (2 * precision * recall) / (precision + recall)
-# chisq.test(c(c(68,12),c(7,64)))
 ```
@@
 This is not surprising -- it is unlikely that any complex cultural phenomenon could be reduced to a small number of cross-cultural predictors.
+
+We can put a p value on these by using a Chi-squared test:
+
+```{r }
+chisq.test(c(c(68,12),c(7,64))) %>% 
+  tidy() %>% 
+  kable('simple', digits = 2)
+```
+
+The model is above chance in predicting its training data.
 
@@
 f2 = (2 * precision2 * recall2) / (precision2 + recall2)
-chisq.test(c(c(264,113),c(87,194)))
+
 ```
@@
 However, the model is still more accurate than it would be by chance.
+
+We can put a p value on these as well by using a Chi-squared test:
+
+```{r }
+chisq.test(c(c(264,113),c(87,194))) %>% 
+  tidy() %>% 
+  kable('simple', digits = 2)
+```
+
+The model is above chance in predicting its test data.
 
@@
 
-```{r fig.width=10,fig.height=15}
+```{r fig.width=10,fig.height=16}
 d_out = ld %>% 
@@
   mutate(
-    possession_trance_present = case_when(
+    `possession trance` = case_when(
       code %in% c(3,4,6,7) ~ 'present',
@@
       ) %>% 
-  select(soc_id,possession_trance_present)
+  select(soc_id,`possession trance`)
 
@@
   mutate(
-    predictor = fct_reorder(var_name, -as.double(recoded_value)),
+    recoded_value2 = case_when(
+      var_title == 'Inheritance rule for real property (land)' ~ as.double(code),
+      var_title != 'Inheritance rule for real property (land)' ~ as.double(recoded_value)
+    ),
+    predictor = fct_reorder(var_name, -recoded_value2),
+    inv_value = -as.integer(recoded_value),
     var_title = fct_relevel(var_title, vars2$Name)
@@
 plots = d_vis %>%
-  count(var_title,predictor,possession_trance_present) %>% 
-  # pivot_wider(names_from = possession_trance_present, values_from = n, values_fill = 0) %>% 
-  # mutate(log_odds_pt = log((`TRUE` + 1) / (`FALSE` + 1)) ) %>% 
+  count(var_title,predictor,`possession trance`) %>% 
   group_by(var_title) %>% 
@@
   mutate(
-    plot = map2(var_title, data, ~ ggplot(.y, aes(predictor,n,fill = possession_trance_present)) +
+    plot = map2(var_title, data, ~ 
+                  ggplot(.y, aes(predictor,n,fill = `possession trance`)) +
                   geom_col(position = 'dodge') +
@@
                   theme(axis.title=element_blank()) +
-                  guides(fill = 'none') +
+                  # guides(fill = 'none') +
                   scale_fill_brewer(palette = "Paired") +
@@
 
-wrap_plots(plots, ncol = 2) + plot_annotation(caption = 'Dark blue: possession trance present. Light blue: absent.')
+wrap_plots(plots, ncol = 2) + plot_layout(guides = "collect") & theme(legend.position = 'bottom')
 ```
+
+Let's step back for a second.
+How do we know that these effects are present beyond cultural autocorrelation?
+Our tree model was fit on the Standard Cross-Cultural Sample, which attempts to control for cultural autocorrelation.
+Another thing we can do is fit a model on the Ethnographic Atlas data but account, to some extent, for autocorrelation.
+This is difficult.
+We can take the seven best predictors from our tree model.
+These already account for over 75% of the explained variation.
+Then, we can take the observations from the Atlas that have data for all seven variables.
+This is 283/658 societies that have data on possession trance in the first place.
+We can then fit a Bayesian hierarchical logistic regression model with a grouping factor for geographic region and for language family.
+There is a trade-off here.
+If we include more predictor variables, we have fewer societies with complete data.
+If we use more complex measures to account for autocorrelation, we will eventually overfit the model.
+This implementation is one of the less suboptimal ones.
+
+In the resulting fit, we see that six of the seven predictors have a robust effect on the outcome, with a 95% credible interval that excludes zero.
+All of the predictors are ordered categories and sometimes the effect is linear, sometimes it's quadratic, or cubic.
+The exception is premarital sexual norms for girls (EA078).
+
+![Regression fig](figures/glm_res.png "Regression coefficients with 90% and 50% credible intervals")
+
+Is this because EA078 is a false positive, once we consider autocorrelation in a different way?
+It is hard to say: Maybe the effect is more robust across all societies, but these will not have data for some or all of the other predictors.
+Clearly, there are a lot of analytic decisions to make here and there is no optimal solution.
+Since the main hypothesis of this paper touches on social rigidity more generally, questions about one predictor will not necessary sink the entire project.
 
@@
 
-We select three phylogenies, Austronesian, Afro-Asiatic, and Atlantic-Congo.
-These are robust phylogenies with a large number of societies in each.
-They include societies that have been discussed by the trance and possesion literature (@blust2000rat, @greenbaum1973, @boddy1989wombs, @lewis1991introduction).
-
+We select two phylogenies, Austronesian and Bantu.
 We use BayesTraits to assess the co-evolution of predictors and outcome (@pagel2007bayestraits).
@@
 
-Six dependent and six independent models are fit, for the relationship of (i) possession trance and the presence of slavery, (ii) possession trance and nuclear/extended family organisation across the three phylogenies.
+Four dependent and four independent models are fit, for the relationship of (i) possession trance and the presence of slavery, (ii) possession trance and nuclear/extended family organisation across the three phylogenies.
 
-Phylogenies come from D-Place.
+Phylogenies come from D-Place which stores these for the original publications.
 Societies are matched to phylogenetic trees by language ID-s.
-One language is used by more than one society in Afro-Asiatic and in Atlantic-Congo respectively, these are removed.
 Societies that have no data on both traits are pruned from the phylogenetic trees.
@@
 A higher value for a Bayes Factor indicates that the more complex model provides a better fit and so its complexity is justified.
-The Bayes Factors for the six model comparisons can be seen below.
+The Bayes Factors for the four model comparisons can be seen below.
 
@@
 
-The results strongly suggest that possession trance and slavery as well as nuclear family organisation developed independently across the three phylogenies.
-One apparent exception is slavery and possession trance in Afro-Asiatic languages, which of course cover the famed description of possession trance practices in the Sudan and the Horn of Africa.
-But a Bayes Factor of 4 is not substantial.
+The results mostly suggest that possession trance and slavery as well as nuclear family organisation developed independently across the two phylogenies.
 Overall, there is little evidence here for trait co-evolution.

### 2023-05-14 model_bayestraits.R

@@ -13,10 +13,20 @@ library(magrittr)
 dw = read_tsv('data/dat_wide.tsv')
 dl = read_tsv('data/dat_long.tsv')
 
-# trees from d-place
-tr_aut = read.nexus("https://raw.githubusercontent.com/D-PLACE/dplace-data/master/phylogenies/glottolog_aust1307/summary.trees")
-tr_afra = read.nexus("https://raw.githubusercontent.com/D-PLACE/dplace-data/master/phylogenies/glottolog_afro1255/summary.trees")
-tr_atla = read.nexus("https://raw.githubusercontent.com/D-PLACE/dplace-data/master/phylogenies/glottolog_atla1278/summary.trees")
+# trees from d-place but please cite original papers
+tr_aut = read.nexus("data/aut.bttrees")
+tr_atlc = read.nexus("data/bantu.bttrees")
+
+# taxa
+tax_aut = read_csv('data/austronesian_taxa.csv') %>% 
+  select(taxon,glottocode)
+tax_atlc = read_csv('data/bantu_taxa.csv') %>% 
+  select(taxon,glottocode)
+
+# tr_aut$tip.label %in% tax_aut$taxon
+# tax_aut$taxon %in% tr_aut$tip.label
+# tr_atlc$tip.label %in% tax_atlc$taxon
+# tax_atlc$taxon %in% tr_atlc$tip.label
 
 # -- formatting -- #
 
@@ -35,119 +45,70 @@ dw2 = dl %>%
 
 d_aut = dw2 %>% 
   filter(family == 'Austronesian') %>% 
+  left_join(tax_aut) %>% 
+  filter(!is.na(taxon)) %>% 
   as.data.frame()
-d_afr = dw2 %>% 
-  filter(family == 'Afro-Asiatic',glottocode != 'kano1249') %>% 
-  as.data.frame()
-d_atl = dw2 %>% 
-  filter(family == 'Atlantic-Congo',glottocode != 'boro1274') %>% 
+d_atlc = dw2 %>% 
+  filter(family == 'Atlantic-Congo') %>% 
+  left_join(tax_atlc) %>% 
+  filter(!is.na(taxon)) %>% 
   as.data.frame()
 
-rownames(d_aut) = d_aut$glottocode
-rownames(d_afr) = d_afr$glottocode
-rownames(d_atl) = d_atl$glottocode
+rownames(d_aut) = d_aut$taxon
+rownames(d_atlc) = d_atlc$taxon
+
 
 # -- bayestraits: out -- #
 
[…]

### 2023-05-14 model_glm.R

@@ -1,0 +1,67 @@
+################################################
+# glm
+################################################
+
+set.seed(1337)
+
+# -- header -- #
+
+setwd('~/Github/Racz2023trance/')
+
+library(tidyverse)
+library(magrittr)
+library(glue)
+library(rstanarm) 
+
+# -- read-in -- #
+
+# data long
+dl = read_tsv('data/dat_long.tsv')
+# data wide
+dw = read_tsv('data/dat_wide.tsv')
+# important vars according to gbm
+vars = read_tsv('models/varimp11.tsv')
+
+# -- wrangling -- #
+
+# vars
+vars %<>% 
+  mutate(cumsump = cumsum(percentage)) %>%
+  filter(cumsump < .8) %>%
+  pull(variable)
+# grouping factors
+dl %<>% 
+  distinct(soc_id,family,region)
+# sort out factor order, add grouping factors
+dw %<>% 
+  select(all_of(c('soc_id','possession_trance_present',vars))) %>% 
+  left_join(dl) %>% 
+  mutate(across(starts_with('EA'), ~ factor(.x, ordered = T)))
+for(i in 3:9){
+  print(nrow(na.omit(dw[,1:i])))
+}
+
+dw2 = na.omit(dw)
+
+# -- model -- #
+
+fit1 = stan_glmer(possession_trance_present ~ 1 + 
+                    EA070_Slavery_type +
+                    EA008_Domestic_organization +
+                    EA078_Norms_of_premarital_sexual_behavior_of_girls +
+                    EA066_Class_differentiation_primary +
+                    EA033_Jurisdictional_hierarchy_beyond_local_community +
+                    EA030_Settlement_patterns +
+                    EA034_Religion_high_gods +
+                    (1|family) +
+                    (1|region), family = binomial, data = dw2, cores = 8, chains = 8, iter = 2000)
+
+plot(fit1, regex_pars = 'EA')
+ggsave('figures/glm_res.png', width = 8, height = 6)
+plot(fit1, "dens_overlay", pars = "(Intercept)")
+plot(fit1, 'rhat')
+broom.mixed::tidy(fit1, conf.int = T, conf.level = .95) %>% 
+  write_tsv('models/glm_estimates.tsv')
+
+# -- pred -- #
+

### 2023-05-14 Readme.Rmd
new final draft 2

@@
 The `helper` file creates the data files.
-The `model` files run the Gradient Boosting Model and the Bayestraits model, respectively.
+The `model` files run the Gradient Boosting Model, the Generalised Linear Model, and the Bayestraits model, respectively.
 
@@
 
-Here is a map of trance and possession phenomena.
+Here is a map of trance and possession phenomena in the Atlas.
 
-```{r fig.width=8, fig.height=11}
+```{r possession_trance_map, fig.width=8, fig.height=11}
 lld = ld %>% 
@@
 
-```{r }
+```{r table_1}
 trance = sccs %>% 
@@
 
-```{r }
+```{r table_2}
 trance %>% 
@@
 
-```{r eval = F}
+```{r table_3, eval = F}
 pred_table = ld %>% 
@@
 
-```{r }
+```{r table_3_ctd}
 pred_table = read_tsv('data/pred_table_added.tsv')
@@
 
-```{r }
+```{r corrplot}
 cord = ld %>% 
@@
 Others, like `EA053` (Sex differences in animal husbandry) or `EA043` (Major type of descent) can be carved up into categories: These are *one-hot encoded*.
+So, for `EA043`, this means that the category originally has four levels, such as “patrilineal”, “matrilineal”, “cognatic”, and “other”, and it will be converted into three categories, which are `true` or `false` for patrilineal, matrilineal, and cognatic, respectively.
+If all three categories are false, the category has the fourth level, which i “other”.
 For details, see the Appendix.
@@
 
-```{r }
+```{r table_4}
 
@@
 
-```{r }
+```{r table_4_ctd}
 chisq.test(c(c(68,12),c(7,64))) %>% 
@@
 
-```{r }
+```{r table_5}
 
@@
 
-```{r }
+```{r table_5_ctd}
 chisq.test(c(c(264,113),c(87,194))) %>% 
@@
 
-```{r }
+```{r table_6}
 vars = ld %>% 
@@
 
-```{r fig.width=10,fig.height=16}
+```{r bigplot, fig.width=10,fig.height=16}
 d_out = ld %>% 
@@
 
+### 7. Validating the EA results
+
 Let's step back for a second.
@@
 All of the predictors are ordered categories and sometimes the effect is linear, sometimes it's quadratic, or cubic.
-The exception is premarital sexual norms for girls (EA078).
+There is one predictor that was deemed important by the boosting model but doesn't do a lot here.
+This is premarital sexual norms for girls (EA078).
 
@@
 
-## 7. Did these traits develop independently?
+If anything, the results are promising.
+Using one set of data and one approach, we identified a set of relevant predictors.
+Using a different (larger) set of data and another approach, we confirmed that most of these seem to be useful.
+
+### 8. Did these traits develop independently?
 
@@
 
-We select two phylogenies, Austronesian and Bantu.
+We select two phylogenies, Austronesian (@gray_language_2009) and Bantu (@grollemund_bantu_2015).
 We use BayesTraits to assess the co-evolution of predictors and outcome (@pagel2007bayestraits).
@@
 
-bayes factor
-```{r }
+```{r table_7}
 #write_sheet(bf, ss = 'https://docs.google.com/spreadsheets/d/1X7ZCnw6LVhjLIy4SiSeK3iZ6o1S144iHMv4DqOaCDEU/edit#gid=0', sheet = 'Table 7')
@@
 
-## 8. Summary
+### 9. Summary
 
@@
 
-## 9. Limitations
+## 10. Limitations
 
-The cross-cultural model is fit on the Standard Cross-Cultural Sample which does not minimise the phylogenetic signal completely (see [here](https://excd.org/2018/03/01/stats-corner-is-the-standard-cross-cultural-sample-really-standard/)).
-The data in the Ethnographic Atlas cannot be independently verified.
+The cross-cultural model is fit on the Standard Cross-Cultural Sample which does not minimise the phylogenetic signal completely (@eff_does_2004).
+The data in the Ethnographic Atlas cannot be independently verified, though it has some external validity (@bahrami-rad_tabulated_2021).
+Using a hierarchical model, as we did here, goes some way in reducing the phylogenetic signal, but this could be improved in the longer term (@claessens_non-independence_2022).
 The outcome variable and the predictors are selected and defined with the existing literature in mind.
@@
 
-```{r }
+```{r app_table_vars}
 kable(vars, 'simple')
@@
 
-```{r }
+```{r app_table_coding}
 coding_table = ld %>% 
@@
 
-```{r, echo=T}
+```{r app_table_hyp, echo=T}
 gbm_params1 = list(

### 2023-05-14 model_bayestraits.R
new final draft 3

@@ -1,3 +1,9 @@
+################################################
+# bayestraits helper
+# pracz
+################################################
+
+
 setwd('~/Github/Racz2023trance/')
 library(tidyverse)
 library(glue)
@@ -6,6 +12,20 @@ library(ape)
 library(ggtree)
 library(bayestraitr) # sam passmore's bayestraits helper library
 library(magrittr)
+
+# -- fun -- #
+
+# run bayestraits in the shell using parameters from the runner table return marg lik
+runBSTR = function(run = 'run1', variable = 'sl', family = 'atl', type = 'Independent'){
+  run_string = glue('./Bayestraits/BayesTraitsV4 Bayestraits/{run}/{type}/{family}_{variable}.bttrees Bayestraits/{run}/{type}/{family}_{variable}.btdata < Bayestraits/{type}_runner.txt')
+  out_string = glue('Bayestraits/{run}/{type}/{family}_{variable}.btdata.Stones.txt')
+  
+  system(run_string)
+  stones = bt_read.stones(out_string)
+  marg_lik = stones$marginal_likelihood
+  return(marg_lik)
+}
+
 
 # -- in -- #
 
@@ -17,7 +37,7 @@ dl = read_tsv('data/dat_long.tsv')
 tr_aut = read.nexus("data/aut.bttrees")
 tr_atlc = read.nexus("data/bantu.bttrees")
 
-# taxa
+# taxa from d-place
 tax_aut = read_csv('data/austronesian_taxa.csv') %>% 
   select(taxon,glottocode)
 tax_atlc = read_csv('data/bantu_taxa.csv') %>% 
@@ -30,12 +50,14 @@ tax_atlc = read_csv('data/bantu_taxa.csv') %>%
 
 # -- formatting -- #
 
+# turning preds binary
 dw %<>% 
   mutate(
     slavery_present = EA070_Slavery_type == "l_1",
     domestic_organisation_nuclear = EA008_Domestic_organization %in% c('l_1','l_2')
   )
 
+# converting
 dw2 = dl %>% 
   distinct(soc_id,society,glottocode,family) %>% 
   left_join(dw) %>% 
@@ -43,11 +65,13 @@ dw2 = dl %>%
   mutate_if(is.logical,as.double) %>% 
   mutate_if(is.double,as.character)
 
+# grabbing relevant data
 d_aut = dw2 %>% 
   filter(family == 'Austronesian') %>% 
[…]

### 2023-05-14 model_gbm.R

@@ -1,5 +1,6 @@
 ################################################
 # gbm
+# pracz
 ################################################
 
 set.seed(1337)

### 2023-05-14 model_glm.R

@@ -1,5 +1,6 @@
 ################################################
 # glm
+# pracz
 ################################################
 
 set.seed(1337)
@@ -32,7 +33,8 @@ vars %<>%
 # grouping factors
 dl %<>% 
   distinct(soc_id,family,region)
-# sort out factor order, add grouping factors
+
+# sort out factor order, drop non-complete cases, add grouping factors
 dw %<>% 
   select(all_of(c('soc_id','possession_trance_present',vars))) %>% 
   left_join(dl) %>% 
@@ -56,12 +58,11 @@ fit1 = stan_glmer(possession_trance_present ~ 1 +
                     (1|family) +
                     (1|region), family = binomial, data = dw2, cores = 8, chains = 8, iter = 2000)
 
+# -- save everything -- #
+
 plot(fit1, regex_pars = 'EA')
 ggsave('figures/glm_res.png', width = 8, height = 6)
 plot(fit1, "dens_overlay", pars = "(Intercept)")
 plot(fit1, 'rhat')
 broom.mixed::tidy(fit1, conf.int = T, conf.level = .95) %>% 
   write_tsv('models/glm_estimates.tsv')
-
-# -- pred -- #
-
